---
title: "Cell Ranger Output QC"
author: "Esther Zhou"
date: "2026-04-01"
output: html_document
---


```{r message=FALSE, warning=FALSE}
library(Seurat)
library(Matrix)
library(pheatmap)
library(hdf5r)
library(dplyr)
library(tidyr)
library(openxlsx)
library(ggplot2)
library(ggrepel)
library(writexl)
library(edgeR)
library(readxl)
```

#Load the datasets
```{r}
data.seurat <- readRDS("/work/users/x/c/xczhou/Chlamydia/processed/data_seurat.rds")
mouse_data <- readRDS("/work/users/x/c/xczhou/Chlamydia/processed/mouse_data.rds")
sample_info <- readRDS("/work/users/x/c/xczhou/Chlamydia/processed/sample_info.rds")
```

#QC heatmap
sample vs sample correlation
```{r}
# Filter genes
keep <- rowSums(mouse_data) > 0
mouse_data <- mouse_data[keep, ]

#Library size normalization and Log2 transformation 
log_cpm <- cpm(mouse_data, log = TRUE, prior.count = 1)

# Correlation matrices
cor_matp <- cor(log_cpm, method = "pearson")
cor_matsp <- cor(log_cpm, method = "spearman")

rownames(sample_info) <- sample_info$sample
# Annotation
annotation_col <- sample_info[colnames(mouse_data), c("group", "day"), drop = FALSE]
annotation_row <- annotation_col

# Make day a factor in the order i want
annotation_col$day <- factor(annotation_col$day, levels = c(-2, 4, 6, 11))
annotation_row$day <- factor(annotation_row$day, levels = c(-2, 4, 6, 11))

# Manual annotation colors
ann_colors <- list(
  group = c(
    Naive = "#0072B2",
    Vaccinated = "#D55E00"
  ),
  day = c(
    "-2" = "#f2f0f7",
    "4"  = "#6a51a3",
    "6"  = "#9e9ac8",
    "11" = "#cbc9e2"
  )
)

#Plot Heatmaps
png("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/pearson_heatmap.png",
    width = 3600, height = 3200, res = 400)
pheatmap(
  cor_matp,
  annotation_col = annotation_col,
  annotation_row = annotation_row,
  annotation_colors = ann_colors,
  border_color = NA,
  main = "Pearson Correlation of Mouse Gene Expression",
  fontsize = 10,
  fontsize_row = 6,
  fontsize_col = 6
)
dev.off()

png("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/spearman_heatmap.png",
    width = 3600, height = 3200, res = 400)
pheatmap(cor_matsp,
         annotation_col = annotation_col,
         annotation_row = annotation_row,
         annotation_colors = ann_colors,
         border_color = NA,
         main = "Spearman Correlation of Mouse Gene Expression",
         fontsize = 10,
         fontsize_row = 6,
         fontsize_col = 6)
dev.off()

#Investigate low correlation samples
colSums(mouse_data)
sort(colSums(mouse_data))

```

#Excluding M52_D4
```{r}
#Excluding M52_D4
mouse_data_no_M52D4 <- mouse_data[, colnames(mouse_data) != "M52_D4"]
sample_info_no_M52D4 <- sample_info[colnames(mouse_data_no_M52D4), ]

log_cpm_no <- cpm(mouse_data_no_M52D4, log = TRUE, prior.count = 1)

# Correlation matrices
cor_matp_no  <- cor(log_cpm_no, method = "pearson")
cor_matsp_no <- cor(log_cpm_no, method = "spearman")

# Annotation data
annotation_col_no <- sample_info_no_M52D4[, c("group", "day"), drop = FALSE]
annotation_row_no <- annotation_col_no

# Make day a factor in the order you want
annotation_col_no$day <- factor(annotation_col_no$day, levels = c(-2, 4, 6, 11))
annotation_row_no$day <- factor(annotation_row_no$day, levels = c(-2, 4, 6, 11))

# Annotation colors
ann_colors <- list(
  group = c(
    Naive = "#0072B2",
    Vaccinated = "#D55E00"
  ),
  day = c(
    "-2" = "#f2f0f7",
    "4"  = "#6a51a3",
    "6"  = "#9e9ac8",
    "11" = "#cbc9e2"
  )
)

# Pearson heatmap
pheatmap(
  cor_matp_no,
  annotation_col = annotation_col_no,
  annotation_row = annotation_row_no,
  annotation_colors = ann_colors,
  border_color = NA,
  main = "Pearson Correlation (M52_D4 removed)",
  show_rownames = TRUE,
  show_colnames = FALSE
)
```


#Excluding M52_D4 and M52_D6

```{r}

#Excluding M52_D4 and M52_D6
mouse_data_no_M52D4D6 <- mouse_data[, !(colnames(mouse_data) %in% c("M52_D4", "M52_D6"))]
sample_info_no_M52D4D6 <- sample_info[colnames(mouse_data_no_M52D4D6), ]

log_cpm_filtered <- cpm(mouse_data_no_M52D4D6, log = TRUE, prior.count = 1)

cor_matp_filtered <- cor(log_cpm_filtered, method = "pearson")
cor_matsp_filtered <- cor(log_cpm_filtered, method = "spearman")

annotation_col_filtered <- sample_info_no_M52D4D6[, c("group", "day"), drop = FALSE]
annotation_row_filtered <- annotation_col_filtered

# Make day a factor in the order you want
annotation_col_filtered$day <- factor(annotation_col_filtered$day, levels = c(-2, 4, 6, 11))
annotation_row_filtered$day <- factor(annotation_row_filtered$day, levels = c(-2, 4, 6, 11))


png("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/pearson_heatmap_d4d6removed.png",
    width = 4800, height = 4200, res = 400)
pheatmap(
  cor_matp_filtered,
  annotation_col = annotation_col_filtered,
  annotation_row = annotation_row_filtered,
  annotation_colors = ann_colors,
  border_color = NA,
  main = "Pearson Correlation (M52_D4 D6 removed)",
  show_rownames = TRUE,
  show_colnames = TRUE
)

dev.off()
```


#Import from Deseq2 filtered.rmd first
#DESeq2 sample sample correlation
```{r}
rld <- rlog(dds_mouse, blind = TRUE)
rld_mat <- assay(rld)

# Sample-sample correlations
rld_corp <- cor(rld_mat, method = "pearson")
rld_cors <- cor(rld_mat, method = "spearman")

# Make sure sample_info rownames match rld_mat columns
rownames(sample_info) <- sample_info$sample

# Annotation data
annotation_col <- sample_info[colnames(rld), c("group", "day"), drop = FALSE]

# Clearer annotation colors
ann_colors <- list(
  group = c(
    Naive = "#0072B2",
    Vaccinated = "orange"
  ),
  day = c(
    "-2" = "#f2f0f7",
    "4"  = "#6a51a3",
    "6"  = "#9e9ac8",
    "11" = "#cbc9e2"
  )
)

# Plot heatmap
heat1_annotation <- pheatmap(
  rld_corp,
  annotation_col = annotation_col,
  annotation_row = annotation_col,   # add this for consistency
  annotation_colors = ann_colors,
  main = "Sample Correlation (rlog-transformed counts) - Pearson",
  show_rownames = FALSE,
  show_colnames = TRUE,
  fontsize = 12
)

heat1_annotation <- pheatmap(
  rld_cors,
  annotation_col = annotation_col,
  annotation_row = annotation_col,   # add this for consistency
  annotation_colors = ann_colors,
  main = "Sample Correlation (rlog-transformed counts) - Spearman",
  show_rownames = FALSE,
  show_colnames = TRUE,
  fontsize = 12
)

ggsave(paste0("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/heatmap_rlog_filtered.tiff"),
       heat1_annotation,
       device = "tiff",
       width =12,
       dpi = 300)


pheatmap(rld_cor, cluster_rows=FALSE, cluster_cols=FALSE,
         main="rlog correlation (no clustering)")
```




#Filter out M52_D4
```{r}
mouse_data <- mouse_data[, colnames(mouse_data) != "M52_D4"]
sample_info <- sample_info[rownames(sample_info) != "M52_D4", ]
data.seurat <- subset(
  data.seurat,
  cells = setdiff(colnames(data.seurat), "M52_D4")
)
```

# Highly expressed Mouse RNA
I computed the mean raw expression for each gene across samples within each day.
```{r}
mouse.exprs <- t(mouse_data) %>% as.data.frame()
mouse.exprs.log2 <- log2(mouse.exprs + 1)

mouse.exprs$day <- sample_info$day
mouse.exprs$group <- sample_info$group

mouse.exprs.log2$day <- sample_info$day
mouse.exprs.log2$group <- sample_info$group

#Top genes
#take all samples in that day, compute mean expression for each gene
#raw
top_genes <- mouse.exprs %>%
  group_by(day) %>%
  summarise(across(-group, \(x) mean(x, na.rm = TRUE))) %>% 
  pivot_longer(-day, names_to = "gene", values_to = "avg_expression") %>%  
  group_by(day) %>%
  arrange(desc(avg_expression)) %>%  
  mutate(label_high = ifelse(avg_expression > 15000, toupper(gene), ""))

top20_genes <- top_genes %>%
  group_by(day) %>%
  slice_head(n = 20)
openxlsx::write.xlsx(top20_genes, "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/top20_genes.xlsx")

#log2
top_genes_log2 <- mouse.exprs.log2 %>%
  group_by(day) %>%
  summarise(across(-group, \(x) mean(x, na.rm = TRUE))) %>% 
  pivot_longer(-day, names_to = "gene", values_to = "avg_expression") %>%  
  group_by(day) %>%
  arrange(desc(avg_expression)) %>%  
  mutate(label_high = ifelse(avg_expression > log2(15000 + 1), toupper(gene), ""))

top20_genes_log2 <- top_genes_log2 %>%
  group_by(day) %>%
  slice_head(n = 20)
openxlsx::write.xlsx(top20_genes_log2,  "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/top20_genes_log2.xlsx")


#Average log2 transformed expression of mouse RNA across all samples measured at 4 timepoints. Selected highly expressed genes are labeled.
#raw
p1 <- ggplot(top_genes, 
             aes(x = factor(day, levels = c(-2, 4, 6, 11)), y = avg_expression)) +
  geom_jitter(width = 0.25, size = 3, alpha = 0.3, color = "grey50") +
  geom_text_repel(aes(label = label_high),
            size = 5,
    box.padding = unit(0.2, "lines"), 
    point.padding = unit(0.2, "lines"),
    fontface = "italic",
    color = "black",
    max.overlaps = Inf,
    force = 4, 
    direction = "both") +
  labs(y = "Average Expression",
       x = "Days post-inoculation (dpi)") +
  theme(axis.text.x = element_text(hjust = 1,size=14, color="black"), 
        axis.text.y = element_text(size = 14, color = "black"),
        axis.title = element_text(size = 16), 
        legend.position = "none",
         panel.background = element_rect(fill = "white"),
          plot.background = element_rect(fill = "white"),
        axis.line.x = element_line(color = "black"),
          axis.line.y = element_line(color = "black"))


ggsave(paste0("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/gene_exprs.tiff"),
       p1,
       device = "tiff",     # Specify the file type
       width = 10,           # Width of the plot (in inches)
       height = 7,          # Height of the plot (in inches)
       dpi = 300)

#log2
p2 <- ggplot(top_genes_log2, 
             aes(x = factor(day, levels = c(-2, 4, 6, 11)), y = avg_expression)) +
  geom_jitter(width = 0.25, size = 1.5, alpha = 0.3, color = "grey50") +
  geom_text_repel(aes(label = label_high),
            size = 5,
    box.padding = unit(0.2, "lines"), 
    point.padding = unit(0.2, "lines"),
    fontface = "italic",
    color = "black",
    max.overlaps = Inf,
    force = 4, 
    direction = "both") +
  labs(y = "Average Log2 Expression",
       x = "Days post-inoculation (dpi)") +
  theme(axis.text.x = element_text(hjust = 1,size=14, color="black"), 
        axis.text.y = element_text(size = 14, color = "black"),
        axis.title = element_text(size = 16), 
        legend.position = "none",
         panel.background = element_rect(fill = "white"),
          plot.background = element_rect(fill = "white"),
        axis.line.x = element_line(color = "black"),
          axis.line.y = element_line(color = "black"))


ggsave(paste0("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/gene_exprs_log(2+1).tiff"),
       p2,
       device = "tiff",     # Specify the file type
       width = 10,           # Width of the plot (in inches)
       height = 7,          # Height of the plot (in inches)
       dpi = 300)
 
```


#Total Chlamydia Reads Detected
```{r}
#Chlamydia-specific probes detected in RNA-seq libraries
chlam_data <- as.matrix(data.seurat[["ChlamydiaRNA"]]$counts)
colSums(chlam_data)

#total reads per gene
gene_total <- rowSums(chlam_data)
gene_total

#number of mice with detection
mice_detected <- rowSums(chlam_data > 0)
mice_detected

#Table 1
result_table <- data.frame(
  Gene = rownames(chlam_data),
  Total_Reads = rowSums(chlam_data),
  Mice_Detected = rowSums(chlam_data > 0)
)

result_table

#Two summaries within each group
chlam_naive <- chlam_data[, sample_info$group == "Naive", drop = FALSE]
chlam_vacc <- chlam_data[, sample_info$group == "Vaccinated", drop = FALSE]

chlam_table <- data.frame(
  Gene = rownames(chlam_data),
  
  Total_Read_Count = rowSums(chlam_data),
  Total_Mice_Detected = paste0(rowSums(chlam_data > 0), "/", ncol(chlam_data)),
  
  Naive_Read_Count = rowSums(chlam_naive),
  Naive_Mice_Detected = paste0(rowSums(chlam_naive > 0), "/", ncol(chlam_naive)),
  
  Vaccinated_Read_Count = rowSums(chlam_vacc),
  Vaccinated_Mice_Detected = paste0(rowSums(chlam_vacc > 0), "/", ncol(chlam_vacc)),
  
  stringsAsFactors = FALSE
)

chlam_table <- chlam_table[order(-chlam_table$Total_Read_Count), ]
chlam_table


#Separated by days
sample_info$subgroup <- paste(sample_info$group, sample_info$day_label)

sample_info$subgroup <- factor(sample_info$subgroup)
sample_info$subgroup <- factor(
  sample_info$subgroup,
  levels = c(
    "Naive D-2",
    "Naive D4",
    "Naive D6",
    "Naive D11",
    "Vaccinated D-2",
    "Vaccinated D4",
    "Vaccinated D6",
    "Vaccinated D11"
  )
)

#extracts the ordered levels of the factor
subgroups <- levels(sample_info$subgroup)

chlam_day_table <- data.frame(Gene = rownames(chlam_data), stringsAsFactors = FALSE)

for (sg in subgroups) {
  idx <- sample_info$subgroup == sg
  
  sub_mat <- chlam_data[, idx, drop = FALSE]
  
  chlam_day_table[[paste0(sg, " Read Count")]] <- rowSums(sub_mat)
  chlam_day_table[[paste0(sg, " # Mice")]] <- paste0(rowSums(sub_mat > 0), "/", ncol(sub_mat))
}

chlam_day_table$Gene <- sub("^Chlamydia_", "", chlam_day_table$Gene)
chlam_day_table <- chlam_day_table[order(-rowSums(chlam_data)), ]

print(chlam_day_table, row.names = FALSE)


#Long Table
subgroups <- expand.grid(
  Group = c("Naive", "Vaccinated"),
  Day = c("-2", "4", "6", "11"),
  stringsAsFactors = FALSE
)

chlam_day_long <- data.frame()

for (i in seq_len(nrow(subgroups))) {
  g <- subgroups$Group[i]
  d <- subgroups$Day[i]
  
  idx <- sample_info$group == g & sample_info$day == d
  sub_mat <- chlam_data[, idx, drop = FALSE]
  
  tmp <- data.frame(
    Gene = rownames(chlam_data),
    Group = g,
    Day = d,
    Read_Count = rowSums(sub_mat),
    Mice_Detected = paste0(rowSums(sub_mat > 0), "/", ncol(sub_mat)),
    stringsAsFactors = FALSE
  )
  
  chlam_day_long <- rbind(chlam_day_long, tmp)
}

chlam_day_long$Gene <- sub("^Chlamydia_", "", chlam_day_long$Gene)
chlam_day_long$Day_Label <- paste0(chlam_day_long$Day, " dpi")
chlam_day_long$Day_Label[chlam_day_long$Day == "-2"] <- "-2 dpi"

chlam_day_long <- chlam_day_long[order(chlam_day_long$Gene,
                                       chlam_day_long$Group,
                                       as.numeric(chlam_day_long$Day)), ]

print(chlam_day_long, row.names = FALSE)

write_xlsx(
  list(
    "Overall_Summary" = chlam_table,
    "By_Group_Day_Wide" = chlam_day_table,
    "By_Group_Day_Long" = chlam_day_long
  ),
  "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Tables/chlamydia_summary_tables.xlsx"
)
```


#Chlamydia Gene Count Figure
```{r}

#Plotting table
chlam_plot_data <- as.data.frame(chlam_data) %>%
  tibble::rownames_to_column("Gene") %>%
  pivot_longer(
    cols = -Gene,
    names_to = "sample",
    values_to = "count"
  ) %>%
  left_join(sample_info, by = "sample") %>%
  mutate(
    Gene = sub("^Chlamydia_", "", Gene),
    day_num = as.numeric(as.character(day))
  )

#create the plotting data
plot_chlam <- chlam_plot_data %>%
  filter(Gene %in% c("16SrRNA", "23SrRNA", "omcB")) %>%
  mutate(
    Gene_group = case_when(
      Gene %in% c("16SrRNA", "23SrRNA") ~ "16SrRNA + 23SrRNA",
      Gene == "omcB" ~ "omcB"
    )
  ) %>%
  group_by(sample, mouse, group, day, day_label, day_num, Gene_group) %>%
  summarise(count = sum(count), .groups = "drop") %>%
  mutate(log10_count = log10(count + 1))

#Figure
ggplot(
  plot_chlam,
  aes(x = day_num, y = log10_count, group = mouse, color = group)
) +
  geom_line(alpha = 0.7, linewidth = 0.7) +
  geom_point(size = 2) +
  facet_wrap(~ Gene_group, scales = "free_y") +
  scale_x_continuous(breaks = c(-2, 4, 6, 11)) +
  labs(
    x = "Days post infection",
    y = expression(log[10]~"(read count + 1)"),
    color = "Group",
    title = "Chlamydia Gene Counts per Mouse Over Time"
  ) +
  theme_bw()

ggsave(
  filename = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/chlamydia_gene_trajectories.png",
  width = 8,
  height = 5,
  dpi = 300
)

```



#Genome equivalents vs rRNA counts
```{r}
# Import genome equivalents
ge_raw <- read_excel(
  "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Genome_Equivalents.xlsx"
)

head(ge_raw)

#Reshape the table
ge_long <- ge_raw %>%
  pivot_longer(
    cols = -Mouse,
    names_to = "day",
    values_to = "genome_equivalent"
  ) %>%
  mutate(
    mouse = as.numeric(Mouse),
    day = as.numeric(day),
    ge_log10 = log10(genome_equivalent + 1)
  ) %>%
  select(mouse, day, genome_equivalent, ge_log10)

#Create total 16S + 23S counts
chlam_long <- as.data.frame(chlam_data) %>%
  tibble::rownames_to_column("Gene") %>%
  pivot_longer(
    cols = -Gene,
    names_to = "sample",
    values_to = "count"
  ) %>%
  left_join(sample_info, by = "sample") %>%
  mutate(
    Gene = sub("^Chlamydia_", "", Gene),
    day = as.numeric(as.character(day))
  )

#Sum 16S + 23S per mouse/day
rna_16s_23s <- chlam_long %>%
  filter(Gene %in% c("16SrRNA", "23SrRNA")) %>%
  group_by(sample, mouse, group, day) %>%
  summarise(
    total_rRNA_count = sum(count),
    .groups = "drop"
  ) %>%
  mutate(
    rna_log10 = log10(total_rRNA_count + 1)
  )

#Merge RNA-seq with genome equivalents
plot_data <- rna_16s_23s %>%
  left_join(ge_long, by = c("mouse", "day"))

plot_data <- plot_data %>%
  filter(!is.na(genome_equivalent))


#Calculate Spearman Overall correlation
#Across all samples, is there an association?
cor_all <- cor.test(
  plot_data$rna_log10,
  plot_data$ge_log10,
  method = "spearman"
)

r_all <- unname(cor_all$estimate)
p_all <- cor_all$p.value

#Group-specific correlations
#Does the relationship differ between Vaccinated vs Naive?
r_group <- plot_data %>%
  group_by(group) %>%
  summarise(
    r = cor(rna_log10, ge_log10, method = "spearman"),
    n = n(),
    .groups = "drop"
  )

r_group

#Label
r_vacc <- r_group$r[r_group$group == "Vaccinated"]
r_naive <- r_group$r[r_group$group == "Naive"]

r_label <- paste0(
  "R = ", round(r_all, 2), "\n",
  "Vaccinated R = ", round(r_vacc, 2), "\n",
  "Naive R = ", round(r_naive, 2)
)

#Plot
p <- ggplot(
  plot_data,
  aes(x = ge_log10, y = rna_log10, color = group, shape = factor(day))
) +
  geom_point(size = 3) +
geom_smooth(
  data = plot_data,
  aes(x = ge_log10, y = rna_log10),
  method = "lm",
  se = FALSE,
  color = "black",
  linetype = "dashed",
  inherit.aes = FALSE
) +
  annotate(
    "text",
    x = min(plot_data$ge_log10, na.rm = TRUE) + 0.2,
    y = max(plot_data$rna_log10, na.rm = TRUE) - 0.2,
    label = r_label,
    hjust = 0,
    vjust = 1,
    size = 4.5
  ) +
  labs(
    x = expression(Log[10]~"Genome Equivalents"),
    y = expression(Log[10]~"rRNA-seq Counts"),
    color = "Group",
    shape = "dpi",
    title = "Chlamydial rRNA Counts vs Genome Equivalents"
  ) +
  scale_color_manual(
    values = c(
      "Naive" = "purple",
      "Vaccinated" = "orange"
    )
) +
  theme_bw()

p

ggsave(
  filename = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/chlamydia_gene_GE.png",
  plot = p,
  width = 8,
  height = 5,
  dpi = 300
)

```



