---
title: "DE Analysis Time Interaction"
author: "Esther Zhou"
date: "2026-05-01"
output: html_document
---


```{r message=FALSE, warning=FALSE}
library(dplyr)
library(broom)
library(tidyr)
library(ggplot2)
library(stringr)
library(matrixStats)
library(knitr)
library(Seurat)
library(corrplot)
library(sparseMatrixStats)
library(SummarizedExperiment)
library(pheatmap)
library(corrplot)
library(ggrepel)
library(DESeq2)
library(EnhancedVolcano)
library(tibble)
library(patchwork)

table_dir <- "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Tables"

dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
```



#Load the datasets
```{r message=FALSE, warning=FALSE}
dds_mouse <- readRDS("/work/users/x/c/xczhou/Chlamydia/processed/dds_mouse.rds")

#Excluding M52_D4
dds_mouse_filtered <- dds_mouse[, !(colnames(dds_mouse) %in% c("M52_D4"))]

# Set reference levels
dds_mouse_filtered$group <- factor(dds_mouse_filtered$group,
                                   levels = c("Naive", "Vaccinated"))

dds_mouse_filtered$day <- factor(dds_mouse_filtered$day,
                                 levels = c("-2", "4", "6", "11"))

```

#Volcano plot
```{r}
plot_DE_adj <- function(All.DE, name){
  
 All.DE <- All.DE %>%
    filter(!is.na(padj), padj < 0.05) %>%
    mutate(
      keys = case_when(
        log2FoldChange < -1 ~ "purple",
        log2FoldChange > 1  ~ "orange",
        TRUE ~ "black"))
 
  # Adaptive labeling for large gene sets
  if (nrow(All.DE) <= 12) {
    # Few genes - label all
    label_genes <- All.DE$gene
    max_overlaps <- Inf
  } else if (nrow(All.DE) <= 100) {
    # Medium - label most extreme
    label_genes <- All.DE %>%
      filter(abs(log2FoldChange) > 2 | padj < 0.01) %>%
      pull(gene)
    max_overlaps <- 20}
  
  keyvals <- All.DE$keys
  names(keyvals)[keyvals == "orange"] <- "Positive interaction"
  names(keyvals)[keyvals == "purple"] <- "Negative interaction"
  names(keyvals)[keyvals == "black"]  <- "Significant"
  
  All.DE$log10_padj <- -log10(All.DE$padj)
  
  # Then, get the max value and add 0.5
  max_y <- max(-log10(All.DE$padj), na.rm = TRUE) + 0.3
  
  DEplot <- EnhancedVolcano(All.DE,
                            lab = All.DE$gene,
                            selectLab = label_genes,
                            x = 'log2FoldChange',
                            y = 'padj',
                            pCutoff = 0.05,
                            FCcutoff = 1,
                            colCustom = keyvals,
                            title = name, 
                            drawConnectors = TRUE, 
                            widthConnectors = 0.5,
                            max.overlaps = 20,
                            subtitle = "",
                            colAlpha = 0.7) +
    labs(x = expression(log[2]~"fold change"),
         y = expression(-log[10]~"(adj p)")) +
    theme(legend.position = "none",
          plot.title = element_text(
            hjust = 0.5, 
            size = 16,
            face = "bold", 
            margin = margin(b = -10)
          ))+
    scale_y_continuous(limits = c(1, max_y))
  
  ggsave(paste0("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/",name,"_DE_interaction.png"), plot = DEplot, dpi = 300, width = 5.3, height = 5)
  
  return(DEplot)
}
```

#Line plot individual
```{r}
plot_interaction_line <- function(dds, gene, name = NULL, day_levels = NULL,
                                  outdir = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/") {

  if (is.null(name)) {
    name <- paste0(gene, "_interaction_line")
  }
  
  plot_data <- plotCounts(
    dds,
    gene = gene,
    intgroup = c("group", "day"),
    normalized = TRUE,
    returnData = TRUE
  )
    # Guard against zeros on log scale
  plot_data$count <- plot_data$count + 1
  
  p <- ggplot(plot_data, aes(x = day, y = count, color = group, group = group)) +
    geom_point(size = 3, alpha = 0.8,
               position = position_jitter(width = 0.08, height = 0)) +
    stat_summary(fun = median, geom = "line", linewidth = 1) +
    stat_summary(fun = median, geom = "point", size = 4) +
    scale_color_manual(
      values = c(
        "Naive" = "purple",
        "Vaccinated" = "orange"
      )
    ) +
    scale_y_log10() +
    theme_bw() +
    labs(
      title = paste("Interaction pattern:", gene),
      x = "Day",
      y = "Normalized count + 1 (log scale)"
    )
  
  ggsave(
    paste0(outdir, name, ".png"),
    plot = p,
    dpi = 300,
    width = 5.3,
    height = 4.5
  )
  
  return(p)
}
```

#Line plot pdf
```{r}
plot_interaction_pdf <- function(dds, gene_list, pdf_name, 
                                  ncol = 4, nrow = 6,
                                  day_levels = NULL,
                                  outdir = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/") {
  
  # Build all plots
  plots <- lapply(gene_list, function(gene) {
    
    plot_data <- plotCounts(
      dds,
      gene = gene,
      intgroup = c("group", "day"),
      normalized = TRUE,
      returnData = TRUE
    )
    
    plot_data$count <- plot_data$count + 1
    
    if (!is.null(day_levels)) {
      plot_data$day <- factor(plot_data$day, levels = day_levels)
    }
    
    ggplot(plot_data, aes(x = day, y = count, color = group, group = group)) +
      geom_point(size = 2, alpha = 0.8,
                 position = position_jitter(width = 0.08, height = 0)) +
      stat_summary(fun = median, geom = "line", linewidth = 0.8) +
      stat_summary(fun = median, geom = "point", size = 3) +
      scale_color_manual(values = c("Naive" = "purple", "Vaccinated" = "orange")) +
      scale_y_log10() +
      theme_bw(base_size = 9) +
      labs(title = gene, x = "Day", y = "Count + 1 (log)") +
      theme(legend.position = "bottom",
            legend.text = element_text(size = 7),
            legend.title = element_blank(),
            plot.title = element_text(face = "bold", size = 9))
  })
  
  # Calculate number of pages needed
  plots_per_page <- ncol * nrow
  n_pages <- ceiling(length(plots) / plots_per_page)
  
  pdf(paste0(outdir, pdf_name, ".pdf"), width = 4 * ncol, height = 3.5 * nrow)
  
  for (i in seq_len(n_pages)) {
    start <- (i - 1) * plots_per_page + 1
    end   <- min(i * plots_per_page, length(plots))
    page_plots <- plots[start:end]
    
    # Fill remaining slots on last page with blank plots
    if (length(page_plots) < plots_per_page) {
      blanks <- plots_per_page - length(page_plots)
      page_plots <- c(page_plots, replicate(blanks, ggplot() + theme_void(), simplify = FALSE))
    }
    
    print(wrap_plots(page_plots, ncol = ncol, nrow = nrow))
  }
  
  dev.off()
  message("Saved: ", outdir, pdf_name, ".pdf")
}
```


# Model 1: group-by-time interaction model

#Baseline and day 4 DESeq
```{r message=FALSE, warning=FALSE}
#filter dataset
dds_d4 <- dds_mouse_filtered[, dds_mouse_filtered$day %in% c("-2", "4")]

# Drop unused day levels
dds_d4$day <- droplevels(dds_d4$day)
dds_d4$group <- droplevels(dds_d4$group)

# Fit interaction model with LRT vs reduced model
#does adding the interaction term significantly improve the model fit?
design(dds_d4) <- ~ group + day + group:day
dds_d4 <- dds_d4[rowSums(counts(dds_d4)) >= 10, ]
dds_int_d4 <- DESeq(dds_d4, test = "LRT", reduced = ~ group + day)


resultsNames(dds_int_d4)

# Extract results (use 0.05 as the FDR threshold)
res_int_d4 <- results(dds_int_d4, alpha = 0.05)

res_int_d4_df <- as.data.frame(res_int_d4) %>%
  rownames_to_column("gene")

```

#Baseline and day 4 significant interaction genes Plots
```{r}
# filter by padj < 0.05
sig_int_d4 <- res_int_d4_df[
  !is.na(res_int_d4_df$padj) &
    res_int_d4_df$padj < 0.05,]

# sort by adjusted p-value
sig_int_d4 <- sig_int_d4[order(sig_int_d4$log2FoldChange), ]

# export selected columns
sig_int_d4_export <- sig_int_d4[, c("gene", "log2FoldChange", "padj")]
rownames(sig_int_d4_export) <- NULL

write.csv(
  sig_int_d4_export,
  file = file.path(table_dir, "Day4_interaction_significant_genes.csv"),
  row.names = FALSE
)

#volcano plot
plot_DE_adj(
  All.DE = res_int_d4_df,
  name = "Day4_interaction"
)

for (g in sig_int_d4$gene) {
  print(plot_interaction_line(
    dds = dds_int_d4,
    gene = g,
    day_levels = c("-2", "4"),
    name = paste0("Day4_", g, "_interaction_line")
  ))
}

plot_interaction_pdf(
  dds = dds_int_d4,
  gene_list = sig_int_d4$gene,
  pdf_name = "Day4_interaction_lines",
  ncol = 4, nrow = 6,
  day_levels = c("-2", "4")
)
###############Not Using#############################
#MA Plot
plotMA(res_int_d4, ylim = c(-5, 5),
       main = "Day 4 interaction: Vaccinated vs Naive change from baseline")

#heatmap
ord <- order(dds_int_d4$group, dds_int_d4$day)
dds_int_d4_ord <- dds_int_d4[, ord]

plot_interaction_heatmap(
  dds = dds_int_d4_ord,
  gene_list = sig_int_d4,
  name = "Day4_interaction_heatmap_ordered",
  top_n = 30,
  intgroup = c("group", "day"),
  transform_method = "rlog",
  cluster_cols = FALSE
)
```

#Day 6
```{r message=FALSE, warning=FALSE}
#filter dataset
dds_d6 <- dds_mouse_filtered[, dds_mouse_filtered$day %in% c("-2", "6")]

# Drop unused day levels
dds_d6$day <- droplevels(dds_d6$day)
dds_d6$group <- droplevels(dds_d6$group)

# Check levels immediately after dropping
levels(dds_d6$day)
levels(dds_d6$group) 

# Fit interaction model with LRT vs reduced model
design(dds_d6) <- ~ group + day + group:day
dds_d6 <- dds_d6[rowSums(counts(dds_d6)) >= 10, ]
dds_int_d6 <- DESeq(dds_d6, test = "LRT", reduced = ~ group + day)

resultsNames(dds_int_d6)

#Extract the interaction results
res_int_d6 <- results(dds_int_d6, alpha = 0.05)

res_int_d6_df <- as.data.frame(res_int_d6) %>%
  rownames_to_column("gene")

```

#Baseline and day 6 significant interaction genes Plots
```{r}
# filter by padj < 0.05
sig_int_d6 <- res_int_d6_df[
  !is.na(res_int_d6_df$padj) &
    res_int_d6_df$padj < 0.05,]

#sorts the significant interaction gene table by adjusted p-value
sig_int_d6 <- sig_int_d6[order(sig_int_d6$log2FoldChange), ]

sig_int_d6_export <- sig_int_d6[, c("gene", "log2FoldChange", "padj")]
rownames(sig_int_d6_export) <- NULL

write.csv(
  sig_int_d6_export,
  file = file.path(table_dir, "Day6_interaction_significant_genes.csv"),
  row.names = FALSE
)


#volcano plot
plot_DE_adj(
  All.DE = res_int_d6_df,
  name = "Day6_interaction"
)

#line plots
for (g in sig_int_d6$gene) {
  print(plot_interaction_line(
    dds = dds_int_d6,
    gene = g,
    day_levels = c("-2", "6"),
    name = paste0("Day6_", g, "_interaction_line")
  ))
}


plot_interaction_pdf(
  dds = dds_int_d6,
  gene_list = sig_int_d6$gene,
  pdf_name = "Day6_interaction_lines",
  ncol = 4, nrow = 6,
  day_levels = c("-2", "6")
)

```

#Day 11
```{r}
#filter dataset
dds_d11 <- dds_mouse_filtered[, dds_mouse_filtered$day %in% c("-2", "11")]

# Drop unused day levels
dds_d11$day <- droplevels(dds_d11$day)
dds_d11$group <- droplevels(dds_d11$group)

# Check levels immediately after dropping
levels(dds_d11$day)
levels(dds_d11$group) 

#fit the interaction model
# Fit interaction model with LRT vs reduced model
design(dds_d11) <- ~ group + day + group:day
dds_d11 <- dds_d11[rowSums(counts(dds_d11)) >= 10, ]
dds_int_d11 <- DESeq(dds_d11, test = "LRT", reduced = ~ group + day)
resultsNames(dds_int_d11)

# Extract LRT interaction results
res_int_d11 <- results(dds_int_d11, alpha = 0.05)
res_int_d11_df <- as.data.frame(res_int_d11) %>%
  rownames_to_column("gene")
```

#Baseline and day 11 significant interaction genes Plots
```{r}
# filter by padj < 0.05
sig_int_d11 <- res_int_d11_df[
  !is.na(res_int_d11_df$padj) &
    res_int_d11_df$padj < 0.05 ,]

# sort significant interaction gene table by log2FC
sig_int_d11 <- sig_int_d11[order(sig_int_d11$log2FoldChange), ]

sig_int_d11_export <- sig_int_d11[, c("gene", "log2FoldChange", "padj")]
rownames(sig_int_d11_export) <- NULL

write.csv(
  sig_int_d11_export,
  file = file.path(table_dir, "Day11_interaction_significant_genes.csv"),
  row.names = FALSE
)


#volcano plot
plot_DE_adj(
  All.DE = res_int_d11_df,
  name = "Day11_interaction"
)

#line plots
for (g in sig_int_d11$gene) {
  print(plot_interaction_line(
    dds = dds_int_d11,
    gene = g,
    day_levels = c("-2", "11"),
    name = paste0("Day11_", g, "_interaction_line")
  ))
}


plot_interaction_pdf(
  dds = dds_int_d11,
  gene_list = sig_int_d11$gene,
  pdf_name = "Day11_interaction_lines",
  ncol = 4, nrow = 6,
  day_levels = c("-2", "11")
)

```


#Check for overlap
```{r}
genes_d4  <- sig_int_d4$gene
genes_d6  <- sig_int_d6$gene
genes_d11 <- sig_int_d11$gene

overlap_summary <- data.frame(
  comparison = c("Day4 & Day6", "Day4 & Day11", "Day6 & Day11", "All three days"),
  n_overlap = c(
    length(intersect(genes_d4, genes_d6)),
    length(intersect(genes_d4, genes_d11)),
    length(intersect(genes_d6, genes_d11)),
    length(Reduce(intersect, list(genes_d4, genes_d6, genes_d11)))
  ),
  genes = c(
    paste(intersect(genes_d4, genes_d6), collapse = ", "),
    paste(intersect(genes_d4, genes_d11), collapse = ", "),
    paste(intersect(genes_d6, genes_d11), collapse = ", "),
    paste(Reduce(intersect, list(genes_d4, genes_d6, genes_d11)), collapse = ", ")
  )
)

overlap_summary
```


#Heat map of the significant genes
```{r}
library(DESeq2)
library(pheatmap)

# 1. Union of significant interaction genes
# unique() ensures Clic6 (present in both D6 and D11) is not duplicated → 38 genes total
interaction_genes <- unique(c(
  sig_int_d4$gene,
  sig_int_d6$gene,
  sig_int_d11$gene
))

# 2. Transform counts
# rlog() stabilizes variance across genes (better than raw counts for visualization)
# blind = FALSE means it uses the experimental design to inform the transformation
rld <- rlog(dds_mouse_filtered, blind = FALSE)
mat <- assay(rld) # extract the transformed count matrix (genes x samples)

# 3. Keep matched genes only
matched_genes <- interaction_genes[interaction_genes %in% rownames(mat)]

# subset to 38 genes x all samples
mat_int <- mat[matched_genes, , drop = FALSE]

# 4. Build annotation directly from mat_int columns
# pulling "group" (Naive/Vaccinated) and "day" (-2/4/6/11) for each sample
anno_col <- as.data.frame(colData(dds_mouse_filtered)[, c("group", "day")])
anno_col <- anno_col[colnames(mat_int), , drop = FALSE]   # ensure same sample order as matrix

# 5. Set factor levels so days are always ordered -2 → 4 → 6 → 11 (not alphabetically)
day_levels <- c("-2", "4", "6", "11")
anno_col$day <- factor(
  as.character(anno_col$day),
  levels = day_levels
)

anno_col$group <- factor(as.character(anno_col$group))  # Naive / Vaccinated

# 6. Build column order manually, day by day
# We need this because cluster_cols = FALSE, so pheatmap will use exactly the column order we give it — we want: Day-2 | Day4 | Day6 | Day11
# within each day: Naive first, then Vaccinated

#Create empty containers
mat_blocks <- list()
anno_blocks <- list()
i <- 1

#Loop through days in the chosen order
for (d in day_levels) {
  
  # Find all samples belonging to this day
  samples_d <- rownames(anno_col)[!is.na(anno_col$day) & anno_col$day == d]
  samples_d <- samples_d[samples_d %in% colnames(mat_int)]
  
  cat("Day:", d, " | Number of samples:", length(samples_d), "\n")
  
  # Skip the day if there are no samples
  if (length(samples_d) == 0) next

# Pull out the expression matrix for this day only
  mat_d <- mat_int[, samples_d, drop = FALSE]

  #Cluster samples within this day
  if (length(samples_d) > 1) {
    #only for clustering columns within each day (temporary)
    mat_d_scaled <- t(scale(t(mat_d)))
    mat_d_scaled[is.na(mat_d_scaled)] <- 0
  
  hc_d <- hclust(dist(t(mat_d_scaled)))
  samples_d <- samples_d[hc_d$order]
}
  
  # Store the ordered submatrix and annotation for this day
  mat_blocks[[i]] <- mat_int[, samples_d, drop = FALSE]
  anno_blocks[[i]] <- anno_col[samples_d, , drop = FALSE]
  
  i <- i + 1
}

# 7. Combine all day blocks side by side into the final ordered matrix
mat_ordered <- do.call(cbind, mat_blocks)
anno_ordered <- do.call(rbind, anno_blocks)

# 8. Safety check: column names of matrix must exactly match row names of annotation
stopifnot(all(colnames(mat_ordered) == rownames(anno_ordered)))

# 9. Calculate where to draw vertical gap lines between day blocks
# cumsum gives the column index at the END of each day block
# we drop the last value because we don't need a gap after the final block
block_sizes <- sapply(mat_blocks, ncol)
gaps_col <- cumsum(block_sizes)
gaps_col <- gaps_col[-length(gaps_col)]

# 10. Define annotation bar colors
annotation_colors <- list(
  group = c(
    Naive = "purple",
    Vaccinated = "orange"
  ),
  day = c(
    "-2" = "#CCCCCC",
    "4"  = "#2171b5",
    "6"  = "#6baed6",
    "11" = "#9ecae1"
  )
)

# 11. Draw the heatmap
# scale = "row"      → z-score each gene across all samples (so color = relative expression)
# cluster_rows = TRUE → genes with similar patterns across samples group together
# cluster_cols = FALSE→ columns stay in the exact order we built above (Day x Group)
# gaps_col           → draws vertical lines between day blocks

#Cap row-scaled heatmap color
my_breaks <- seq(-3, 3, length.out = 101)


pheatmap(
  mat_ordered,
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = anno_ordered,
  annotation_colors = annotation_colors,
  gaps_col = gaps_col,
  show_colnames = FALSE,
  fontsize_row = 11,
  main = "Significant Group-by-Time Interaction Genes",
  breaks = my_breaks
)
```


#Trajectory plots

```{r}
plot_trajectory_pdf <- function(dds, gene_list, title, pdf_name,
                                 ncol = 4, nrow = 6,
                                 day_levels = c("-2", "4", "6", "11"),
                                 outdir = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/") {
  
  # Build one plot per gene
  plots <- lapply(gene_list, function(g) {
    
    plot_data <- plotCounts(
      dds,
      gene = g,
      intgroup = c("group", "day"),
      normalized = TRUE,
      returnData = TRUE
    ) %>%
      mutate(count = count + 1)
    
    plot_data$day <- factor(plot_data$day, levels = day_levels)
    
    ggplot(plot_data, aes(x = day, y = count, color = group, group = group)) +
      geom_point(size = 2, alpha = 0.6,
                 position = position_jitter(width = 0.08, height = 0)) +
      stat_summary(fun = median, geom = "line", linewidth = 0.9) +
      stat_summary(fun = median, geom = "point", size = 3) +
      scale_color_manual(values = c("Naive" = "purple", "Vaccinated" = "orange")) +
      scale_y_log10() +
      theme_bw(base_size = 9) +
      labs(title = g, x = "Day", y = "Count + 1 (log)") +
      theme(legend.position = "bottom",
            legend.title = element_blank(),
            plot.title = element_text(face = "bold", size = 9))
  })
  
  # Paginate
  plots_per_page <- ncol * nrow
  n_pages <- ceiling(length(plots) / plots_per_page)
  
  pdf(paste0(outdir, pdf_name, ".pdf"), width = 4 * ncol, height = 3.5 * nrow)
  
  for (i in seq_len(n_pages)) {
    start <- (i - 1) * plots_per_page + 1
    end   <- min(i * plots_per_page, length(plots))
    page_plots <- plots[start:end]
    
    # Fill blank slots on last page
    if (length(page_plots) < plots_per_page) {
      blanks <- plots_per_page - length(page_plots)
      page_plots <- c(page_plots, replicate(blanks, ggplot() + theme_void(), simplify = FALSE))
    }
    
    print(wrap_plots(page_plots, ncol = ncol, nrow = nrow))
  }
  
  dev.off()
  message("Saved: ", outdir, pdf_name, ".pdf")
}

# Day 4 - 3 genes
plot_trajectory_pdf(
  dds = dds_mouse_filtered,
  gene_list = sig_int_d4$gene,
  pdf_name = "Day4_sig_alldays_trajectory",
  ncol = 3, nrow = 1
)

# Day 6 - 24 genes
plot_trajectory_pdf(
  dds = dds_mouse_filtered,
  gene_list = sig_int_d6$gene,
  pdf_name = "Day6_sig_alldays_trajectory",
  ncol = 4, nrow = 6
)

# Day 11 - 12 genes
plot_trajectory_pdf(
  dds = dds_mouse_filtered,
  gene_list = sig_int_d11$gene,
  pdf_name = "Day11_sig_alldays_trajectory",
  ncol = 4, nrow = 3
)
```


#Model 2: Genes that change over time similarly in both groups
Among the remaining genes, which genes change from −2 to day 4 in a similar/shared way across both groups?

#day 4
```{r}
# Remove interaction-significant genes from the object before fitting
dds_main_d4 <- dds_d4[!(rownames(dds_d4) %in% sig_int_d4$gene), ]
design(dds_main_d4) <- ~ group + day
dds_main_d4 <- DESeq(dds_main_d4)
resultsNames(dds_main_d4)

#extract the day effect
res_time_d4 <- results(
  dds_main_d4,
  name = "day_4_vs_.2",
  alpha = 0.05
)

res_time_d4_df <- as.data.frame(res_time_d4) %>%
  rownames_to_column("gene")

#Volcano plot
plot_DE(
  All.DE = res_time_d4_df,
  name = "Day4_shared_time_effect"
)

## Export uses stricter filtered table
shared_sig_d4_export <- res_time_d4_df %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) > 1) %>%
  arrange(padj) %>%
  select(gene, log2FoldChange, padj)

write.csv(shared_sig_d4_export,
          file = file.path(table_dir, "Day4_shared_time_significant_genes.csv"),
          row.names = FALSE)

#heatmap
shared_sig_d4 <- res_time_d4_df %>%
  filter(!is.na(padj), padj < 0.05)

top_shared_up_d4 <- shared_sig_d4 %>%
  filter(log2FoldChange > 0) %>%
  arrange(padj) %>%
  slice_head(n = 15)

top_shared_down_d4 <- shared_sig_d4 %>%
  filter(log2FoldChange < 0) %>%
  arrange(padj) %>%
  slice_head(n = 15)

top_shared_d4 <- bind_rows(top_shared_up_d4, top_shared_down_d4)


ord <- order(dds_main_d4$group, dds_main_d4$day)
dds_main_d4_ord <- dds_main_d4[, ord]

# Heatmap of top shared Day 4 response genes
plot_interaction_heatmap(
  dds = dds_main_d4_ord,
  gene_list = top_shared_d4,
  name = "Day4_shared_time_top30_heatmap",
  intgroup = c("group", "day"),
  transform_method = "rlog",
  cluster_cols = FALSE
)

```


#day6

```{r}
dds_main_d6 <- dds_d6[!(rownames(dds_d6) %in% sig_int_d6$gene), ]
design(dds_main_d6) <- ~ group + day
dds_main_d6 <- DESeq(dds_main_d6)
resultsNames(dds_main_d6)

#extract the day effect
res_time_d6 <- results(
  dds_main_d6,
  name = "day_6_vs_.2",
  alpha = 0.05
)

res_time_d6_df <- as.data.frame(res_time_d6) %>%
  rownames_to_column("gene")

#Volcano plot
plot_DE(
  All.DE = res_time_d6_df,
  name = "Day6_shared_time_effect"
)

## Export uses stricter filtered table
shared_sig_d6_export <- res_time_d6_df %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) > 1) %>%
  arrange(padj) %>%
  select(gene, log2FoldChange, padj)

write.csv(shared_sig_d6_export,
          file = file.path(table_dir, "Day6_shared_time_significant_genes.csv"),
          row.names = FALSE)


#heatmap
shared_sig_d6 <- res_time_d6_df %>%
  filter(!is.na(padj), padj < 0.05)

top_shared_up_d6 <- shared_sig_d6 %>%
  filter(log2FoldChange > 0) %>%
  arrange(padj) %>%
  slice_head(n = 15)

top_shared_down_d6 <- shared_sig_d6 %>%
  filter(log2FoldChange < 0) %>%
  arrange(padj) %>%
  slice_head(n = 15)

top_shared_d6 <- bind_rows(top_shared_up_d6, top_shared_down_d6)


ord <- order(dds_main_d6$group, dds_main_d6$day)
dds_main_d6_ord <- dds_main_d6[, ord]

# Heatmap of top shared Day 6 response genes
plot_interaction_heatmap(
  dds = dds_main_d6_ord,
  gene_list = top_shared_d6,
  name = "Day6_shared_time_top30_heatmap",
  intgroup = c("group", "day"),
  transform_method = "rlog",
  cluster_cols = FALSE
)


```


#day11
```{r}
# day11
dds_main_d11 <- dds_d11[!(rownames(dds_d11) %in% sig_int_d11$gene), ]
design(dds_main_d11) <- ~ group + day
dds_main_d11 <- DESeq(dds_main_d11)
resultsNames(dds_main_d11)

# extract the day effect
res_time_d11 <- results(
  dds_main_d11,
  name = "day_11_vs_.2",
  alpha = 0.05
)

res_time_d11_df <- as.data.frame(res_time_d11) %>%
  rownames_to_column("gene")


# Volcano plot
plot_DE(
  All.DE = res_time_d11_df,
  name = "Day11_shared_time_effect"
)

# Export uses stricter filtered table
shared_sig_d11_export <- res_time_d11_df %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) > 1) %>%
  arrange(padj) %>%
  select(gene, log2FoldChange, padj)

write.csv(shared_sig_d11_export,
          file = file.path(table_dir, "Day11_shared_time_significant_genes.csv"),
          row.names = FALSE)

# Heatmap
shared_sig_d11 <- res_time_d11_df %>%
  filter(!is.na(padj), padj < 0.05)

top_shared_up_d11 <- shared_sig_d11 %>%
  filter(log2FoldChange > 0) %>%
  arrange(padj) %>%
  slice_head(n = 15)

top_shared_down_d11 <- shared_sig_d11 %>%
  filter(log2FoldChange < 0) %>%
  arrange(padj) %>%
  slice_head(n = 15)

top_shared_d11 <- bind_rows(top_shared_up_d11, top_shared_down_d11)

ord <- order(dds_main_d11$group, dds_main_d11$day)
dds_main_d11_ord <- dds_main_d11[, ord]

# Heatmap of top shared Day 11 response genes
plot_interaction_heatmap(
  dds = dds_main_d11_ord,
  gene_list = top_shared_d11,
  name = "Day11_shared_time_top30_heatmap",
  intgroup = c("group", "day"),
  transform_method = "rlog",
  cluster_cols = FALSE
)
```



#Heatmap function
```{r}
plot_interaction_heatmap <- function(dds,
                                     gene_list,
                                     name,
                                     intgroup = c("group", "day"),
                                     transform_method = c("rlog", "vst"),
                                     top_n = NULL,
                                     cluster_rows = TRUE,
                                     cluster_cols = FALSE,
                                     scale_rows = TRUE,
                                     outdir = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/") {
  
  transform_method <- match.arg(transform_method)
  
  # If gene_list is a DE result dataframe, extract genes
  if (is.data.frame(gene_list)) {
    if (!"gene" %in% colnames(gene_list)) {
      stop("If gene_list is a dataframe, it must contain a column named 'gene'.")
    }
    
    if (!is.null(top_n) && "padj" %in% colnames(gene_list)) {
      gene_list <- gene_list %>%
        filter(!is.na(padj)) %>%
        arrange(padj) %>%
        slice_head(n = top_n) %>%
        pull(gene)
    } else {
      gene_list <- gene_list$gene
    }
  }
  
  # Keep only genes present in dds
  gene_list <- intersect(gene_list, rownames(dds))
  
  if (length(gene_list) == 0) {
    stop("No genes in gene_list are found in the DESeqDataSet.")
  }
  
  # Transform counts
  if (transform_method == "rlog") {
    trans_obj <- rlog(dds, blind = TRUE)
  } else {
    trans_obj <- vst(dds, blind = TRUE)
  }
  
  mat <- assay(trans_obj)[gene_list, , drop = FALSE]
  
  # Scale each gene across samples
  if (scale_rows) {
    mat <- t(scale(t(mat)))
  }
  
  # Sample annotation
  ann_col <- as.data.frame(colData(dds)[, intgroup, drop = FALSE])
  ann_colors <- list(group = c("Naive" = "purple", "Vaccinated" = "orange"))
  ann_colors <- ann_colors[names(ann_colors) %in% colnames(ann_col)]
  
  # Save heatmap
  pheatmap(
    mat,
    annotation_col = ann_col,
    annotation_colors = ann_colors,
    cluster_rows = cluster_rows,
    cluster_cols = cluster_cols,
    show_rownames = TRUE,
    show_colnames = FALSE,
    fontsize_row = 9,
    main = name,
    filename = file.path(outdir, paste0(name, "_heatmap.png")),
    width = 7,
    height = max(4.5, length(gene_list) * 0.25)
  )
}
```



#heatmap
```{r}
#1. Union all significant genes
shared_sig_d4_padj <- res_time_d4_df %>%
  filter(!is.na(padj), padj < 0.05) %>%
  pull(gene)

shared_sig_d6_padj <- res_time_d6_df %>%
  filter(!is.na(padj), padj < 0.05) %>%
  pull(gene)

shared_sig_d11_padj <- res_time_d11_df %>%
  filter(!is.na(padj), padj < 0.05) %>%
  pull(gene)

shared_genes_all <- unique(c(
  shared_sig_d4_padj,
  shared_sig_d6_padj,
  shared_sig_d11_padj
))

length(shared_genes_all)

#Transform and subset to all shared-time genes
rld <- rlog(dds_mouse_filtered, blind = FALSE)
mat <- assay(rld)

matched_shared_genes <- shared_genes_all[shared_genes_all %in% rownames(mat)]

mat_int <- mat[matched_shared_genes, , drop = FALSE]

dim(mat_int)

#Use the same within-day sample clustering procedure
anno_col <- as.data.frame(colData(dds_mouse_filtered)[, c("group", "day")])
anno_col <- anno_col[colnames(mat_int), , drop = FALSE]

day_levels <- c("-2", "4", "6", "11")

anno_col$day <- factor(as.character(anno_col$day), levels = day_levels)
anno_col$group <- factor(as.character(anno_col$group))

mat_blocks <- list()
anno_blocks <- list()
i <- 1

for (d in day_levels) {
  
  samples_d <- rownames(anno_col)[!is.na(anno_col$day) & anno_col$day == d]
  samples_d <- samples_d[samples_d %in% colnames(mat_int)]
  
  cat("Day:", d, " | Number of samples:", length(samples_d), "\n")
  
  if (length(samples_d) == 0) next
  
  mat_d <- mat_int[, samples_d, drop = FALSE]
  
  if (length(samples_d) > 1) {
    mat_d_scaled <- t(scale(t(mat_d)))
    mat_d_scaled[is.na(mat_d_scaled)] <- 0
    
    hc_d <- hclust(dist(t(mat_d_scaled)))
    samples_d <- samples_d[hc_d$order]
  }
  
  mat_blocks[[i]] <- mat_int[, samples_d, drop = FALSE]
  anno_blocks[[i]] <- anno_col[samples_d, , drop = FALSE]
  
  i <- i + 1
}

mat_ordered <- do.call(cbind, mat_blocks)
anno_ordered <- do.call(rbind, anno_blocks)

stopifnot(all(colnames(mat_ordered) == rownames(anno_ordered)))

block_sizes <- sapply(mat_blocks, ncol)
gaps_col <- cumsum(block_sizes)
gaps_col <- gaps_col[-length(gaps_col)]


#heatmap, use k-means for clustering? (kmeans_k = 10)
annotation_colors <- list(
  group = c(
    Naive = "purple",
    Vaccinated = "orange"
  ),
  day = c(
    "-2" = "#CCCCCC",
    "4"  = "#2171b5",
    "6"  = "#6baed6",
    "11" = "#9ecae1"
  )
)

my_breaks <- seq(-3, 3, length.out = 101)

ph <- pheatmap(
  mat_ordered,
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  treeheight_row = 250,   # increase row dendrogram/linkage size
  treeheight_col = 0,
  annotation_col = anno_ordered,
  annotation_colors = annotation_colors,
  gaps_col = gaps_col,
  show_colnames = FALSE,
  show_rownames = TRUE,
  fontsize_row = 1,
  breaks = my_breaks,
  main = "Shared Time-Effect Genes",
  filename = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Shared_Time_Effect_Genes_heatmap_large.pdf",
  width = 14,
  height = 80
)

#20 Clusters
set.seed(123)

ph <- pheatmap(
  mat_ordered,
  scale = "row",
  kmeans_k = 20,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  treeheight_row = 150,   # increase row dendrogram/linkage size
  treeheight_col = 0,
  annotation_col = anno_ordered,
  annotation_colors = annotation_colors,
  gaps_col = gaps_col,
  show_colnames = FALSE,
  show_rownames = FALSE,
  breaks = my_breaks,
  main = "Shared Time-Effect Genes",
  filename = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Shared_Time_Effect_Genes_heatmap_20.pdf",
  width = 10,
  height = 20
)

gene_clusters <- data.frame(
  gene = names(ph$kmeans$cluster),
  cluster = ph$kmeans$cluster
)

head(gene_clusters)
table(gene_clusters$cluster)
write.csv(
  gene_clusters,
  file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Shared_Time_Effect_Genes_kmeans20_clusters.csv",
  row.names = FALSE
)

```

