---
title: "DESeq D4"
author: "Esther Zhou"
date: "2026-04-03"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(broom)
library(tidyr)
library(ggplot2)
library(stringr)
library(matrixStats)
library(knitr)
library(Seurat)
library(corrplot)
library(sparseMatrixStats)
library(SummarizedExperiment)
library(pheatmap)
library(corrplot)
library(ggrepel)
library(DESeq2)
library(EnhancedVolcano)
```



#Load the datasets
```{r message=FALSE, warning=FALSE}
dds_mouse <- readRDS("/work/users/x/c/xczhou/Chlamydia/processed/dds_mouse.rds")
sample_info <- readRDS("/work/users/x/c/xczhou/Chlamydia/processed/sample_info.rds")
```

#Excluding M52_D4
```{r}
dds_mouse_filtered <- dds_mouse[, !(colnames(dds_mouse) %in% c("M52_D4"))]
```

#Sample-level QC to assess overall similarity between samples
Using PCA and Hierarchical clustering
rlog: 
1. Preprocessing for plotting (PCA, Heatmaps) rather than differential expression testing (use DESeq for testing)
2. It handles the dependence of variance on the mean, improving upon simple transformations.

deseq2::plotPCA is a function in the DESeq2 R package used to visualize sample-to-sample distances using Principal Component Analysis (PCA). It is primarily used for quality control to identify batch effects, outliers, or expected grouping in RNA-seq data,
```{r}
#Run rlog transformation
rld <- rlog(dds_mouse_filtered, blind=TRUE)

# Plot PCA
DESeq2::plotPCA(rld, ntop = 500, intgroup = "group")
DESeq2::plotPCA(rld, ntop = 500, intgroup = "day")

PCA_data <- DESeq2::plotPCA(rld, ntop = 500,intgroup = "group", returnData = TRUE)
PCA_data$day <- colData(rld)$day

#Save plot
p_pca <- ggplot(PCA_data, aes(x = PC1, y = PC2, color = group, shape = factor(day))) +
  geom_point(size = 4) +
  theme_bw()

ggsave(
  filename = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/PCA_plot_filtered.png",
  plot = p_pca,
  width = 6,
  height = 5,
  dpi = 300
)

```


#PCA plot
check clustering by group/day
PCA summarizes global expression variation, where each point represents a sample, and clustering indicates similarity between biological conditions.
```{r}
#PCA plot
p1 <- ggplot(PCA_data,
             aes(x=PC1, y=PC2, color=group, shape=day)) +
        geom_point(aes(shape=factor(day), color=group), size=5)+
  scale_color_manual(values=c("purple", "orange"))+
  labs(color = "Group", x = "PC1", y = "PC2", shape = "dpi") + 
  scale_shape_manual(values = c(15, 16, 17, 18)) + 
  theme(axis.text.x = element_text(), #angle = 30
        panel.background = element_rect(fill = "white"),  # Black background
        plot.background = element_rect(fill = "white"),   # Black plot area
        axis.text = element_text(color = "black", size = 16), # Axis text white, bold, larger
        axis.title = element_text(color = "black", size = 18), # Axis titles white, bold, larger
        legend.position = "bottom",  # Bottom-right corner
        legend.justification = c(1, 0),
        legend.direction = "horizontal",
        legend.spacing.x = unit(0.2, 'cm'),  # Reduce horizontal spacing between items
        legend.spacing.y = unit(0.2, 'cm'),  # Reduce vertical spacing between items
        legend.key.size = unit(0.4, 'cm'),   # Make the size of the legend keys smaller
        legend.margin = margin(2, 2, 2, 2),  
        legend.text = element_text(size = 18 ),  # Increase legend font size face = "bold"
        legend.title = element_text(size = 18, color = "black"),  # Bold legend title
        axis.line.x = element_line(color = "black"),
        axis.line.y = element_line(color = "black"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white", color = NA)) +
  #legend.box = "horizontal") +
  guides(shape = guide_legend(order = 2),
         color = guide_legend(order = 1))

p1

ggsave(
  "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/PCA_plot_filtered.tiff",
  p1,
  device = "tiff",
  width = 12,
  height = 8,   # ← increase this (try 8–10)
  dpi = 300
)



```





#MANOVA
Multivariate Analysis of Variance (MANOVA) is a statistical technique used to compare multiple continuous dependent variables across two or more groups (independent variables) simultaneously. It determines if group means differ significantly by considering interrelationships between outcomes, offering greater power than running multiple separate ANOVAs
Qestion: Do group (Naive vs Vaccinated) and day explain variation in the multivariate outcome (PC1, PC2 together)?
H0: The mean vector (PC1, PC2) is the same across groups
H1: At least one group has a different (PC1, PC2) mean
```{r}
### Testing (PC1 & PC2) ~ Group + gpi
manova_model <- manova(cbind(PC1, PC2) ~ group + day, data = PCA_data)

# Get a summary of the MANOVA model
summary(manova_model)

# The summary of each individual ANOVA for the responses
summary.aov(manova_model)

```
## Interpretation:
Variation along PC1 is driven by time (day)
PC2 captures variation associated with group (strongly) and day (also contributes)


#Making a volcano plot for differential expression results
Keep only genes with pvalue < 0.05
label:the most strongly upregulated genes; the most strongly downregulated genes; the most statistically significant genes
Assign colors to gene: purple = significantly downregulated (log2FC < -1); orange = significantly upregulated (log2FC > 1); black = significant, but fold change is between -1 and 1

compute -log10(padj)


```{r}
#Function to visualize DE
plot_DE <- function(All.DE, name){
  #Keep only genes with pvalue < 0.05
  All.DE <-  All.DE %>% filter(pvalue < 0.05)  %>%
    #Choose which genes get labels
    mutate(label.print = case_when(
      log2FoldChange > quantile(log2FoldChange, 0.975) ~ gene,
      log2FoldChange < quantile(log2FoldChange, 0.025) ~ gene,
      pvalue < quantile(pvalue, 0.05) ~ gene,
      TRUE ~ ""
    ),
    #Assign colors to genes
    keys = case_when(log2FoldChange < -1 & pvalue < 0.05 ~ 'purple',
                     log2FoldChange > 1 & pvalue < 0.05 ~ "orange",
                     TRUE ~ "black"))
  #tells EnhancedVolcano how to name the color categories.
  keyvals <- All.DE$keys
  names(keyvals)[keyvals == 'orange'] <- 'Up-regulated'
  names(keyvals)[keyvals == 'purple'] <- 'Down-regulated'
  names(keyvals)[keyvals == 'black'] <- 'Significant'
  
  All.DE$log10_pvalue <- -log10(All.DE$pvalue)
  # Then, get the max value and add 0.5
  max_y <- max(All.DE$log10_pvalue, na.rm = TRUE) + 0.3
  
  DEplot <- EnhancedVolcano(All.DE,
                            lab = All.DE$gene,
                            selectLab = All.DE$label.print,
                            x = 'log2FoldChange',
                            y = 'pvalue',
                            pCutoff = 0.05,
                            FCcutoff = 1,
                            colCustom = keyvals,
                            title = name, 
                            #legendLabels = c('', ''), 
                            drawConnectors = TRUE, 
                            widthConnectors = 0.5,
                            max.overlaps = 20,
                            subtitle = "",
                            colAlpha = 0.7) +
    theme(legend.position = "none",
          plot.title = element_text(
            hjust = 0.5, 
            size = 16,         # make the title larger
            face = "bold", 
            margin = margin(b = -10)  # reduce space below the title
          ))+
    scale_y_continuous(limits = c(1, max_y))
  
  ggsave(paste0("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/",name,"_DE_filteredM52D4.png"), plot = DEplot, dpi = 300, width = 5.3, height = 5)
  # width = 4.3, height = 5
  
}

plot_DE_adj <- function(All.DE, name){
  
  All.DE <-  All.DE %>% filter(padj < 0.05)  %>%
    mutate(label.print = case_when(
      log2FoldChange > quantile(log2FoldChange, 0.975) ~ gene,
      log2FoldChange < quantile(log2FoldChange, 0.025) ~ gene,
      padj < quantile(padj, 0.05) ~ gene,
      TRUE ~ ""
    ),
    keys = case_when(log2FoldChange < -1 & padj < 0.05 ~ 'purple',
                     log2FoldChange > 1 & padj < 0.05 ~ "orange",
                     TRUE ~ "black"))
  
  keyvals <- All.DE$keys
  
  names(keyvals)[keyvals == 'orange'] <- 'Up-regulated'
  
  names(keyvals)[keyvals == 'purple'] <- 'Down-regulated'
  
  names(keyvals)[keyvals == 'black'] <- 'Significant'
  
  All.DE$log10_padj <- -log10(All.DE$padj)
  
  # Then, get the max value and add 0.5
  max_y <- max(All.DE$log10_padj, na.rm = TRUE) + 0.3
  
  DEplot <- EnhancedVolcano(All.DE,
                            lab = All.DE$gene,
                            selectLab = All.DE$label.print,
                            x = 'log2FoldChange',
                            y = 'padj',
                            pCutoff = 0.05,
                            FCcutoff = 1,
                            colCustom = keyvals,
                            title = name, 
                            #legendLabels = c('', ''), 
                            drawConnectors = TRUE, 
                            widthConnectors = 0.5,
                            max.overlaps = 20,
                            subtitle = "",
                            colAlpha = 0.7) +
    labs(x = expression(log[2]~"fold change"),
         y = expression(-log[10]~"(adj p)")) +
    theme(legend.position = "none",
          plot.title = element_text(
            hjust = 0.5, 
            size = 16,         # make the title larger
            face = "bold", 
            margin = margin(b = -10)  # reduce space below the title
          ))+
    scale_y_continuous(limits = c(1, max_y))
  
  ggsave(paste0("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/",name,"_DE_padj.png"), plot = DEplot, dpi = 300, width = 5.3, height = 5)
  
}


#DE p-adj with less labeled genes
plot_DE_adj_cleaner <- function(All.DE, name){
  
  All.DE <-  All.DE %>% filter(padj < 0.05)  %>%
    mutate(label.print = case_when(
      log2FoldChange > quantile(log2FoldChange, 0.99) ~ gene,
      log2FoldChange < quantile(log2FoldChange, 0.01) ~ gene,
      padj < 1e-5 ~ gene,
      TRUE ~ ""
    ),
    keys = case_when(log2FoldChange < -1 & padj < 0.05 ~ 'purple',
                     log2FoldChange > 1 & padj < 0.05 ~ "orange",
                     TRUE ~ "black"))
  
  keyvals <- All.DE$keys
  
  names(keyvals)[keyvals == 'orange'] <- 'Up-regulated'
  
  names(keyvals)[keyvals == 'purple'] <- 'Down-regulated'
  
  names(keyvals)[keyvals == 'black'] <- 'Significant'
  
  All.DE$log10_padj <- -log10(All.DE$padj)
  
  # Then, get the max value and add 0.5
  max_y <- max(All.DE$log10_padj, na.rm = TRUE) + 0.3
  
  DEplot <- EnhancedVolcano(All.DE,
                            lab = All.DE$gene,
                            selectLab = All.DE$label.print,
                            x = 'log2FoldChange',
                            y = 'padj',
                            pCutoff = 0.05,
                            FCcutoff = 1,
                            colCustom = keyvals,
                            title = name, 
                            drawConnectors = TRUE, 
                            widthConnectors = 0.3,
                            labSize = 3.5,
                            max.overlaps = 20,
                            subtitle = "",
                            colAlpha = 0.7) +
    labs(x = expression(log[2]~"fold change"),
         y = expression(-log[10]~"(adj p)")) +
    theme(legend.position = "none",
          plot.title = element_text(
            hjust = 0.5, 
            size = 16,
            face = "bold", 
            margin = margin(b = -10)  # reduce space below the title
          ))+
    scale_y_continuous(limits = c(1, max_y), expand = c(0,0)) +
    scale_x_continuous(limits=c(-2.8, 2.8), expand = c(0,0))
  
  ggsave(paste0("/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/",name,"_DE_padj_cleaned.png"), plot = DEplot, dpi = 300, width = 5.3, height = 5)
  
}


```

#DE Analysis Day4 w/ filtered data
padj is generated by DESeq2 using the Benjamini–Hochberg (BH) method to control FDR.

lfcSE: This is the uncertainty (standard error) of the log2 fold change.
Without shrinkage:
1. you often see huge log2FC values for low-count genes
2. those are usually false positives / unstable estimates
With shrinkage:
1. volcano plots look cleaner
2. rankings of genes are more reliable
lfcShrink() uses both log2foldchange and lfcse to decide "should i trust this effect size?"

log2FoldChange > 0 means Higher in Vaccinated mice
log2FoldChange < 0 means Higher in Naive mice
```{r}
#Filter data
dds4 <- dds_mouse_filtered[, dds_mouse_filtered$day==4]

# Remove genes with all zero counts in Day 4 samples
dds4 <- dds4[which(rowSums(dds4@assays@data@listData[["counts"]]) != 0),]
dds4

# Set group reference level
dds4$group <- factor(dds4$group, levels=c("Naive","Vaccinated"))
#Model gene expression as a function of group
design(dds4) <- ~ group

dds4 <- DESeq(dds4)

plotDispEsts(dds4)

# Check the coefficients for the comparison
resultsNames(dds4)

# Generate results object, log2FoldChange = Vaccinated − Naive
res_vaccine_naive_day4 <- results(dds4, 
                alpha = 0.05,
               contrast=c("group","Vaccinated","Naive"))


res_vaccine_naive_day4 <- lfcShrink(
  dds = dds4,
  coef = 2,
  type = "apeglm"
)

head(res_vaccine_naive_day4 )

res_vaccine_naive_day4_table <- as.data.frame(res_vaccine_naive_day4)
res_vaccine_naive_day4_table$gene <- rownames(res_vaccine_naive_day4_table)

#Check for baseMean pvalue and padj NAs
cat("baseMean == 0:", sum(res_vaccine_naive_day4_table$baseMean == 0), "\n")
cat("NA pvalue:", sum(is.na(res_vaccine_naive_day4_table$pvalue)), "\n")
cat("NA padj:", sum(is.na(res_vaccine_naive_day4_table$padj)), "\n")

#Remove genes with NA adjusted p-values
res_vaccine_naive_day4_table <- res_vaccine_naive_day4_table %>% filter(!is.na(padj))
#Count significant genes (raw p-value)
sum(res_vaccine_naive_day4_table$pvalue < 0.05)

plot_DE(All.DE = res_vaccine_naive_day4_table,name="Day 4")

write.csv(res_vaccine_naive_day4_table%>%
            #sort all genes from largest to smallest log2FoldChange
            arrange(-log2FoldChange)%>%
            filter(pvalue<0.05 & log2FoldChange > 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day4_up_DE_p_filtered.csv")

write.csv(res_vaccine_naive_day4_table%>%
            arrange(log2FoldChange)%>%
            filter(pvalue<0.05 & log2FoldChange < 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day4_down_DE_p_filtered.csv")


plot_DE_adj(All.DE = res_vaccine_naive_day4_table%>%dplyr::filter(!is.na(padj)),
            name="Day 4 p-adj")

write.csv(res_vaccine_naive_day4_table%>%
            arrange(-log2FoldChange)%>%
            filter(padj<0.05 & log2FoldChange > 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day4_up_DE_padj_filtered.csv")

write.csv(res_vaccine_naive_day4_table%>%
            arrange(log2FoldChange)%>%
            filter(padj<0.05 & log2FoldChange < 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day4_down_DE_padj_filtered.csv")


#Top DE genes heatmap
rld4 <- rlog(dds4, blind = TRUE)
mat <- assay(rld4)

# choose significant & top 50 genes
top_genes <- res_vaccine_naive_day4_table %>%
  filter(padj < 0.05) %>%
  arrange(padj) %>%
  slice_head(n = 50) %>%
  pull(gene)


```


```{r}
#Upregulated in Vaccinated
up_vaccinated <- res_vaccine_naive_day4_table %>%
  filter(padj < 0.05 & log2FoldChange > 1) %>%
  arrange(desc(log2FoldChange))

#Upregulated in Naive
up_naive <- res_vaccine_naive_day4_table %>%
  filter(padj < 0.05 & log2FoldChange < -1) %>%
  arrange(log2FoldChange)
```


#Plotcounts() some of the up and down regulated genes 
normalized counts (adjusted for library size / sequencing depth)
```{r}
plotCounts(dds4, gene="Il17a", intgroup="group")
plotCounts(dds4, gene="Cd3e", intgroup="group")
plotCounts(dds4, gene="S100a8", intgroup="group")
plotCounts(dds4, gene="C2cd4a", intgroup="group")
plotCounts(dds4, gene="Ank", intgroup="group")

```


#Day 6

```{r}
#Filter data
dds6 <- dds_mouse_filtered[, dds_mouse_filtered$day==6]

# Remove genes with all zero counts in Day 6 samples
dds6 <- dds6[which(rowSums(dds6@assays@data@listData[["counts"]]) != 0),]
dds6

# Set group reference level
dds6$group <- factor(dds6$group, levels=c("Naive","Vaccinated"))
#Model gene expression as a function of group
design(dds6) <- ~ group

dds6 <- DESeq(dds6)

plotDispEsts(dds6)

# Check the coefficients for the comparison
resultsNames(dds6)

# Generate results object
res_vaccine_naive_day6 <- results(dds6, 
                alpha = 0.05,
               contrast=c("group","Vaccinated","Naive"))


res_vaccine_naive_day6 <- lfcShrink(dds=dds6 , coef=2, type="apeglm" )

head(res_vaccine_naive_day6 )

res_vaccine_naive_day6_table <- as.data.frame(res_vaccine_naive_day6)
res_vaccine_naive_day6_table$gene <- rownames(res_vaccine_naive_day6_table)

#Check for baseMean pvalue and padj NAs
cat("baseMean == 0:", sum(res_vaccine_naive_day6_table$baseMean == 0), "\n")
cat("NA pvalue:", sum(is.na(res_vaccine_naive_day6_table$pvalue)), "\n")
cat("NA padj:", sum(is.na(res_vaccine_naive_day6_table$padj)), "\n")

#Remove genes with NA adjusted p-values
res_vaccine_naive_day6_table <- res_vaccine_naive_day6_table %>% filter(!is.na(padj))
#Count significant genes (raw p-value)
sum(res_vaccine_naive_day6_table$pvalue < 0.05)

plot_DE(All.DE = res_vaccine_naive_day6_table,name="Day 6")

write.csv(res_vaccine_naive_day6_table%>%
            #sort all genes from largest to smallest log2FoldChange
            arrange(-log2FoldChange)%>%
            filter(pvalue<0.05 & log2FoldChange > 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day6_up_DE_p_filtered.csv")

write.csv(res_vaccine_naive_day6_table%>%
            arrange(log2FoldChange)%>%
            filter(pvalue<0.05 & log2FoldChange < 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day6_down_DE_p_filtered.csv")


plot_DE_adj(All.DE = res_vaccine_naive_day6_table%>%dplyr::filter(!is.na(padj)),
            name="Day 6 p-adj")

write.csv(res_vaccine_naive_day6_table%>%
            arrange(-log2FoldChange)%>%
            filter(padj<0.05 & log2FoldChange > 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day6_up_DE_padj_filtered.csv")

write.csv(res_vaccine_naive_day6_table%>%
            arrange(log2FoldChange)%>%
            filter(padj<0.05 & log2FoldChange < 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day6_down_DE_padj_filtered.csv")


plot_DE_adj_cleaner(All.DE = res_vaccine_naive_day6_table,name="Day 6 p-adj")
```



# Day 11
```{r}
#Filter data
dds11 <- dds_mouse_filtered[, dds_mouse_filtered$day==11]

# Remove genes with all zero counts in Day 11 samples
dds11 <- dds11[which(rowSums(dds11@assays@data@listData[["counts"]]) != 0),]
dds11

# Set group reference level
dds11$group <- factor(dds11$group, levels=c("Naive","Vaccinated"))
#Model gene expression as a function of group
design(dds11) <- ~ group

dds11 <- DESeq(dds11)

plotDispEsts(dds11)

# Check the coefficients for the comparison
resultsNames(dds11)

# Generate results object
res_vaccine_naive_day11 <- results(dds11, 
                alpha = 0.05,
               contrast=c("group","Vaccinated","Naive"))


res_vaccine_naive_day11 <- lfcShrink(dds=dds11 , coef=2, type="apeglm" )

head(res_vaccine_naive_day11)

res_vaccine_naive_day11_table <- as.data.frame(res_vaccine_naive_day11)
res_vaccine_naive_day11_table$gene <- rownames(res_vaccine_naive_day11_table)


#Check for baseMean pvalue and padj NAs
cat("baseMean == 0:", sum(res_vaccine_naive_day11_table$baseMean == 0), "\n")
cat("NA pvalue:", sum(is.na(res_vaccine_naive_day11_table$pvalue)), "\n")
cat("NA padj:", sum(is.na(res_vaccine_naive_day11_table$padj)), "\n")

#Remove genes with NA adjusted p-values
res_vaccine_naive_day11_table <- res_vaccine_naive_day11_table %>% filter(!is.na(padj))
#Count significant genes (raw p-value)
sum(res_vaccine_naive_day11_table$pvalue < 0.05)

plot_DE(All.DE = res_vaccine_naive_day11_table,name="Day 11")

write.csv(res_vaccine_naive_day11_table%>%
            #sort all genes from largest to smallest log2FoldChange
            arrange(-log2FoldChange)%>%
            filter(pvalue<0.05 & log2FoldChange > 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day11_up_DE_p_filtered.csv")

write.csv(res_vaccine_naive_day11_table%>%
            arrange(log2FoldChange)%>%
            filter(pvalue<0.05 & log2FoldChange < 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day11_down_DE_p_filtered.csv")


plot_DE_adj(All.DE = res_vaccine_naive_day11_table%>%dplyr::filter(!is.na(padj)),
            name="Day 11 p-adj")

write.csv(res_vaccine_naive_day11_table%>%
            arrange(-log2FoldChange)%>%
            filter(padj<0.05 & log2FoldChange > 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day11_up_DE_padj_filtered.csv")

write.csv(res_vaccine_naive_day11_table%>%
            arrange(log2FoldChange)%>%
            filter(padj<0.05 & log2FoldChange < 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day11_down_DE_padj_filtered.csv")

```


# Day -2
```{r}
#Filter data
dds0 <- dds_mouse_filtered[, dds_mouse_filtered$day==-2]

# Remove genes with all zero counts in Day -2 samples
dds0 <- dds0[which(rowSums(dds0@assays@data@listData[["counts"]]) != 0),]
dds0

# Set group reference level
dds0$group <- factor(dds0$group, levels=c("Naive","Vaccinated"))
#Model gene expression as a function of group
design(dds0) <- ~ group

dds0 <- DESeq(dds0)

plotDispEsts(dds0)

# Check the coefficients for the comparison
resultsNames(dds0)

# Generate results object
res_vaccine_naive_day0 <- results(dds0, 
                alpha = 0.05,
               contrast=c("group","Vaccinated","Naive"))


res_vaccine_naive_day0 <- lfcShrink(dds=dds0 , coef=2, type="apeglm" )

head(res_vaccine_naive_day0)

res_vaccine_naive_day0_table <- as.data.frame(res_vaccine_naive_day0)
res_vaccine_naive_day0_table$gene <- rownames(res_vaccine_naive_day0_table)


#Check for baseMean pvalue and padj NAs
cat("baseMean == 0:", sum(res_vaccine_naive_day0_table$baseMean == 0), "\n")
cat("NA pvalue:", sum(is.na(res_vaccine_naive_day0_table$pvalue)), "\n")
cat("NA padj:", sum(is.na(res_vaccine_naive_day0_table$padj)), "\n")

#Check distribution of padj
hist(res_vaccine_naive_day0_table$padj,
     breaks = 10,
     xlim = c(0.995, 1),
     col = "grey",
     border = "white",
     main = "",
     xlab = "Adjusted p-value (padj)",
     ylab = "Frequency")

hist(res_vaccine_naive_day0_table$pvalue, 
     breaks = 50,
     main = "Distribution of Raw p-values (Day -2)",
     xlab = "Raw p-value",
     ylab = "Frequency")

#Remove genes with NA adjusted p-values
res_vaccine_naive_day0_table <- res_vaccine_naive_day0_table %>% filter(!is.na(padj))
#Count significant genes (raw p-value)
sum(res_vaccine_naive_day0_table$pvalue < 0.05)

plot_DE(All.DE = res_vaccine_naive_day0_table,name="Day -2")

write.csv(res_vaccine_naive_day0_table%>%
            #sort all genes from largest to smallest log2FoldChange
            arrange(-log2FoldChange)%>%
            filter(pvalue<0.05 & log2FoldChange > 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day-2_up_DE_p_filtered.csv")

write.csv(res_vaccine_naive_day0_table%>%
            arrange(log2FoldChange)%>%
            filter(pvalue<0.05 & log2FoldChange < 0), 
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/Day-2_down_DE_p_filtered.csv")


plot_DE_adj(All.DE = res_vaccine_naive_day0_table%>%dplyr::filter(!is.na(padj)),
            name="Day -2")

nrow(res_vaccine_naive_day0_table)

sum(!is.na(res_vaccine_naive_day0_table$padj))
sum(res_vaccine_naive_day0_table$padj < 0.05, na.rm = TRUE)

summary(res_vaccine_naive_day0_table$padj)
range(res_vaccine_naive_day0_table$log2FoldChange, na.rm = TRUE)

```


#Are Day 4, 6 and 11 biologically consistent?
```{r}
#Number of DE genes across days
sum(res_vaccine_naive_day4_table$padj < 0.05)
sum(res_vaccine_naive_day6_table$padj < 0.05)
sum(res_vaccine_naive_day11_table$padj < 0.05)

#Are key genes consistent?
genes <- c("Il17a","Cd3d","S100a8")

res_vaccine_naive_day4_table[genes, ]
res_vaccine_naive_day6_table[genes, ]
res_vaccine_naive_day11_table[genes, ]

#Direction consistency
intersect(
  res_vaccine_naive_day4_table$gene[res_vaccine_naive_day4_table$padj < 0.05],
  res_vaccine_naive_day6_table$gene[res_vaccine_naive_day6_table$padj < 0.05]
)


```



#Make top 15 tables
```{r}

top15_up_day6 <- res_vaccine_naive_day6_table %>%
  filter(!is.na(padj)) %>%
  arrange(desc(log2FoldChange)) %>%
  slice_head(n = 15) %>%
  select(gene, baseMean, log2FoldChange, pvalue, padj) %>%
  mutate(
    baseMean = round(baseMean, 3),
    log2FoldChange = round(log2FoldChange, 3),
    pvalue = signif(pvalue, 3),
    padj = signif(padj, 3)
  )

rownames(top15_up_day6) <- NULL
top15_up_day6



#Make the LaTex table
library(knitr)
library(kableExtra)

kable(
  top15_up_day6,
  format = "latex",
  booktabs = TRUE,
  caption = "Top 15 upregulated genes in vaccinated mice at Day 6",
  align = c("l", "r", "r", "r", "r")
) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))

```
