---
title: "GSEA"
author: "Esther Zhou"
date: "2026-04-14"
output: html_document
---

```{r setup, include=FALSE}
library(fgsea)
library(DESeq2)
library(msigdbr)
library(dplyr)
library(ggplot2)
library(stringr)
library(clusterProfiler)
library(org.Mm.eg.db)
library(KEGGREST)
library(pheatmap)
library(tidyr)
```

#Load Dataset and exclude M52_D4
```{r}
dds_mouse <- readRDS("/work/users/x/c/xczhou/Chlamydia/processed/dds_mouse.rds")
sample_info <- readRDS("/work/users/x/c/xczhou/Chlamydia/processed/sample_info.rds")
dds_mouse_filtered <- dds_mouse[, !(colnames(dds_mouse) %in% c("M52_D4"))]

```

#Run DESeq2 to get a ranking statistics for day 4, 6, and 11
```{r message=FALSE, warning=FALSE}
dds4 <- dds_mouse_filtered[, dds_mouse_filtered$day==4]

# Remove genes with all zero counts in Day 4 samples
dds4 <- dds4[which(rowSums(dds4@assays@data@listData[["counts"]]) != 0),]
dds4

# Set group reference level
dds4$group <- factor(dds4$group, levels=c("Naive","Vaccinated"))
#Model gene expression as a function of group
design(dds4) <- ~ group
dds4 <- DESeq(dds4)


#Day 6
dds6 <- dds_mouse_filtered[, dds_mouse_filtered$day==6]
dds6 <- dds6[which(rowSums(dds6@assays@data@listData[["counts"]]) != 0),]
dds6$group <- factor(dds6$group, levels=c("Naive","Vaccinated"))
design(dds6) <- ~ group
dds6 <- DESeq(dds6)

#Day 11
dds11 <- dds_mouse_filtered[, dds_mouse_filtered$day==11]
dds11 <- dds11[which(rowSums(dds11@assays@data@listData[["counts"]]) != 0),]
dds11$group <- factor(dds11$group, levels=c("Naive","Vaccinated"))
design(dds11) <- ~ group
dds11 <- DESeq(dds11)

```


#Rank the genes based on their stat day4
```{r}
#extract stat
res_d4 <- results(dds4, contrast = c("group", "Vaccinated", "Naive"))

# ranked list for GSEA
rank_d4 <- res_d4$stat
names(rank_d4) <- rownames(res_d4)

# remove NA and sort decreasing
rank_d4 <- rank_d4[!is.na(rank_d4)]
rank_d4 <- sort(rank_d4, decreasing = TRUE)

#Make sure gene names match MSigDB format
head(rownames(dds4))
#Check for duplicated gene names
sum(duplicated(names(rank_d4)))
#Check for infinite values
sum(!is.finite(rank_d4))
```

#Rank the genes based on their stat day6
```{r}
#extract stat
res_d6 <- results(dds6, contrast = c("group", "Vaccinated", "Naive"))

# ranked list for GSEA
rank_d6 <- res_d6$stat
names(rank_d6) <- rownames(res_d6)

# remove NA and sort decreasing
rank_d6 <- rank_d6[!is.na(rank_d6)]
rank_d6 <- sort(rank_d6, decreasing = TRUE)

#Make sure gene names match MSigDB format
head(rownames(dds6))
#Check for duplicated gene names
sum(duplicated(names(rank_d6)))
#Check for infinite values
sum(!is.finite(rank_d6))
```

#Rank the genes based on their stat day11

```{r}
#extract stat
res_d11 <- results(dds11, contrast = c("group", "Vaccinated", "Naive"))

# ranked list for GSEA
rank_d11 <- res_d11$stat
names(rank_d11) <- rownames(res_d11)

# remove NA and sort decreasing
rank_d11 <- rank_d11[!is.na(rank_d11)]
rank_d11 <- sort(rank_d11, decreasing = TRUE)

#Make sure gene names match MSigDB format
head(rownames(dds11))
#Check for duplicated gene names
sum(duplicated(names(rank_d11)))
#Check for infinite values
sum(!is.finite(rank_d11))
```

#Create KEGG mouse pathway list
```{r}
# Pull KEGG mouse pathway-gene links
kegg_links <- KEGGREST::keggLink("pathway", "mmu")

kegg_df <- data.frame(
  entrez_id  = sub("mmu:", "", names(kegg_links)),
  pathway_id = sub("path:", "", as.character(kegg_links))
)

# Pull KEGG pathway names
kegg_names <- KEGGREST::keggList("pathway", "mmu")

kegg_name_df <- data.frame(
  pathway_id   = sub("path:", "", names(kegg_names)),
  pathway_name = as.character(kegg_names)
)

# Add readable pathway names
kegg_df <- kegg_df %>%
  left_join(kegg_name_df, by = "pathway_id") %>%
  filter(!is.na(pathway_name))

# Build KEGG pathway list
pathways_kegg_entrez <- split(kegg_df$entrez_id, kegg_df$pathway_name)


#Convert ranked vector from SYMBOL to ENTREZ
convert_rank_to_entrez <- function(rank) {
  
  gene_df <- bitr(
    names(rank),
    fromType = "SYMBOL",
    toType   = "ENTREZID",
    OrgDb    = org.Mm.eg.db
  )
  
  gene_df <- gene_df %>%
    mutate(stat = rank[SYMBOL]) %>%
    filter(!is.na(stat), !is.na(ENTREZID)) %>%
    group_by(ENTREZID) %>%
    slice_max(order_by = abs(stat), n = 1, with_ties = FALSE) %>%
    ungroup()
  
  rank_entrez <- gene_df$stat
  names(rank_entrez) <- gene_df$ENTREZID
  
  sort(rank_entrez, decreasing = TRUE)
}

rank_d4_entrez  <- convert_rank_to_entrez(rank_d4)
rank_d6_entrez  <- convert_rank_to_entrez(rank_d6)
rank_d11_entrez <- convert_rank_to_entrez(rank_d11)
```

#Retrieve the collections of genes from MSigDB
```{r}
#Pull the mouse M7 nad M8 gene sets
msig_m7 <- msigdbr(
  db_species = "MM",
  species = "Mus musculus",
  collection = "M7"
)

msig_m8 <- msigdbr(
  db_species = "MM",
  species = "Mus musculus",
  collection = "M8"
)

#Turn them into pathway lists
# .$ means take the data coming from the pipe (%>%) and use it here
pathways_m7 <- msig_m7 %>%
  split(x = .$gene_symbol, f = .$gs_name)
pathways_m8 <- msig_m8 %>%
  split(x = .$gene_symbol, f = .$gs_name)

```

#Reactome gene list
```{r}
reactome_df <- msigdbr(
  species = "Mus musculus",
  collection = "C2",
  subcollection = "CP:REACTOME"
)

pathways_reactome <- reactome_df %>%
  split(x = .$gene_symbol, f = .$gs_name)

```


#GSEA Function
```{r}
run_fgsea_analysis <- function(rank_list,
                               pathways,
                               collection_name,   # e.g. "M7", "M8", "KEGG", "Reactome"
                               day_label,         # e.g. "Day4", "Day6"
                               output_dir,
                               min_size = 10,
                               max_size = 500,
                               padj_cutoff = 0.05,
                               seed = 42) {
  
  # Overlap check
  all_pathway_genes <- unique(unlist(pathways))
  pct_overlap <- round(sum(names(rank_list) %in% all_pathway_genes) / length(rank_list) * 100, 1)
  cat(collection_name, day_label, "- gene overlap:", pct_overlap, "%\n")
  
  # Run fgsea
  set.seed(seed)
  result <- fgsea(
    pathways = pathways,
    stats    = rank_list,
    minSize  = min_size,
    maxSize  = max_size
  ) %>% arrange(padj)
  
  # Summary
  n_sig <- sum(result$padj < padj_cutoff, na.rm = TRUE)
  cat(collection_name, day_label, ":",
      n_sig, "/", nrow(result), "pathways significant (",
      round(n_sig / nrow(result) * 100, 1), "%)\n\n")
  
  # Collapse leadingEdge for export
  result_out <- result %>%
    mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))
  
  result_sig_out <- result_out %>%
    filter(padj < padj_cutoff)
  
  # Write CSVs
  file_prefix <- file.path(output_dir, paste0("fgsea_", tolower(collection_name), "_", tolower(day_label)))
  
# Clean output table
result_out_clean <- result_out %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative")
  ) %>%
  dplyr::select(pathway, direction, NES, padj, size, leadingEdge)

result_sig_out_clean <- result_out_clean %>%
  filter(padj < padj_cutoff)

write.csv(
  result_sig_out_clean,
  file = paste0(file_prefix, "_sig.csv"),
  row.names = FALSE
)

write.csv(
  result_out_clean,
  file = paste0(file_prefix, "_all.csv"),
  row.names = FALSE
)
  
  # Return the raw result (list columns intact) for plotting
  return(result)
}

# ── Top pathway extractor ───────────────────────────────────────────────
get_top_pathways <- function(fgsea_result, n = 10, padj_cutoff = 0.05) {
  
  # Convert data.table to data.frame to avoid dplyr conflicts
  fgsea_result <- as.data.frame(fgsea_result)
  
  sig <- fgsea_result %>%
    filter(padj < padj_cutoff) %>%
    mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))
  
  up <- sig %>%
    filter(NES > 0) %>%
    arrange(desc(NES)) %>%
    dplyr::select(pathway, NES, ES, pval, padj, size, leadingEdge) %>%
    slice_head(n = n)
  
  down <- sig %>%
    filter(NES < 0) %>%
    arrange(NES) %>%
    dplyr::select(pathway, NES, ES, pval, padj, size, leadingEdge) %>%
    slice_head(n = n)
  
  list(up = up, down = down)
}


# ── Plotting function ───────────────────────────────────────────────────
plot_top_pathways <- function(top_pathways,
                               collection_name,
                               day_label,
                               output_dir,
                               plot_type = "both") {  # "bar", "dot", or "both"
  
  # Combine up and down
  plot_df <- bind_rows(top_pathways$up, top_pathways$down) %>%
    mutate(
      direction  = ifelse(NES > 0, "Positive", "Negative"),
      pathway    = reorder(pathway, NES),
      log10padj  = -log10(padj)
    )
  
  if (nrow(plot_df) == 0) {
    cat("No significant pathways for", collection_name, day_label, "- skipping plot\n")
    return(invisible(NULL))
  }
  
  title_base <- paste("Top enriched", collection_name, "pathways at", day_label)
  file_prefix <- file.path(output_dir, paste0("gsea_", tolower(collection_name), "_", tolower(day_label)))
  
  # ── Bar plot ──
  if (plot_type %in% c("bar", "both")) {
    p_bar <- ggplot(plot_df, aes(x = pathway, y = NES, fill = direction)) +
      geom_col() +
      coord_flip() +
      scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
      labs(
        title = title_base,
        x     = "Pathway",
        y     = "Normalized Enrichment Score (NES)",
        fill  = "Direction"
      ) +
      theme_bw() +
      theme(
        plot.title   = element_text(face = "bold", size = 11),
        axis.text.y  = element_text(size = 8),
        legend.position = "bottom"
      )
    
    ggsave(paste0(file_prefix, "_barplot.pdf"), p_bar, width = 10, height = 7)
    ggsave(paste0(file_prefix, "_barplot.png"), p_bar, width = 10, height = 7, dpi = 300)
  }
  
  # ── Dot plot ──
  if (plot_type %in% c("dot", "both")) {
    p_dot <- ggplot(plot_df, aes(x = NES, y = pathway)) +
      geom_point(aes(size = size, color = padj)) +
      scale_color_gradient(low = "darkblue", high = "lightblue") +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      labs(
        title  = title_base,
        x      = "Normalized Enrichment Score (NES)",
        y      = "Pathway",
        size   = "Gene set size",
        color  = "Adjusted p-value"
      ) +
      theme_bw() +
      theme(
        plot.title   = element_text(face = "bold", size = 11),
        axis.text.y  = element_text(size = 8),
        legend.position = "right"
      )
    
    ggsave(paste0(file_prefix, "_dotplot.pdf"), p_dot, width = 10, height = 7)
    ggsave(paste0(file_prefix, "_dotplot.png"), p_dot, width = 10, height = 7, dpi = 300)
  }
  
  cat("Saved plots for", collection_name, day_label, "\n")
  invisible(list(bar = p_bar, dot = p_dot))
}


# ── Heatmap function ────────────────────────────────────────────────────
plot_gsea_heatmap <- function(fgsea_list,
                               collection_name,
                               output_dir,
                               padj_cutoff = 0.05,
                               top_n = 30,
                               value = "NES") {
  
  # Find union of top significant pathways across all days
  top_pathways <- lapply(fgsea_list, function(res) {
    res %>%
      as.data.frame() %>%
      filter(padj < padj_cutoff) %>%
      arrange(padj) %>%
      slice_head(n = top_n) %>%
      pull(pathway)
  })
  selected_pathways <- unique(unlist(top_pathways))
  
  if (length(selected_pathways) == 0) {
    cat("No significant pathways for", collection_name, "- skipping heatmap\n")
    return(invisible(NULL))
  }
  
  # Build long format
  heatmap_df <- lapply(names(fgsea_list), function(day) {
    fgsea_list[[day]] %>%
      as.data.frame() %>%
      filter(pathway %in% selected_pathways) %>%
      mutate(
        day      = day,
        plot_val = if (value == "NES") NES else -log10(padj)
      ) %>%
      dplyr::select(pathway, day, plot_val)
  }) %>%
    bind_rows()
  
  # Reshape to wide format
  mat_df <- reshape(
    heatmap_df,
    idvar     = "pathway",
    timevar   = "day",
    direction = "wide"
  )
  colnames(mat_df) <- sub("plot_val\\.", "", colnames(mat_df))
  
  # Convert to matrix                      ← only ONE matrix block
  mat <- as.matrix(mat_df[, -1])
  rownames(mat) <- mat_df$pathway
  mat[is.na(mat)] <- 0
  
  # Shorten long pathway names
  rownames(mat) <- stringr::str_trunc(rownames(mat), width = 60)
  
# Color scale
  if (value == "NES") {
    max_val       <- max(abs(mat), na.rm = TRUE)
    breaks        <- seq(-max_val, max_val, length.out = 101)
    colors        <- colorRampPalette(c("tomato", "white", "steelblue"))(100)
    legend_breaks <- c(-max_val, 0, max_val)
    legend_labels <- c("Negative", "0", "Positive")
  } else {
    max_val       <- max(mat, na.rm = TRUE)
    breaks        <- seq(0, max_val, length.out = 101)
    colors        <- colorRampPalette(c("white", "darkblue"))(100)
    legend_breaks <- round(seq(0, max_val, length.out = 5), 1)
    legend_labels <- as.character(legend_breaks)
  }
  
  # Shared pheatmap arguments
  heatmap_args <- list(
    mat,
    color         = colors,
    breaks        = breaks,
    cluster_cols  = FALSE,
    cluster_rows  = TRUE,
    fontsize_row  = 8,
    fontsize_col  = 10,
    main          = title,
    border_color  = NA,
    angle_col     = 0,
    legend_breaks = legend_breaks,
    legend_labels = legend_labels
  )
  
  title       <- paste(collection_name, "pathway enrichment across timepoints -", value)
  file_prefix <- file.path(output_dir, paste0("heatmap_", tolower(collection_name), "_", tolower(value)))
  
  # Shared pheatmap arguments
  heatmap_args <- list(
    mat,
    color        = colors,
    breaks       = breaks,
    cluster_cols = FALSE,
    cluster_rows = TRUE,
    fontsize_row = 8,
    fontsize_col = 10,
    main         = title,
    border_color = NA,
    angle_col    = 0,
    legend_breaks = if (value == "NES") c(-max_val, 0, max_val) else waiver(),
    legend_labels = if (value == "NES") c("Negative", "0", "Positive") else waiver()
  )
  
  # Save PDF
  pdf(paste0(file_prefix, ".pdf"), width = 10, height = max(6, length(selected_pathways) * 0.25))
  do.call(pheatmap::pheatmap, heatmap_args)
  dev.off()
  
  # Save PNG
  png(paste0(file_prefix, ".png"), width = 10, height = max(6, length(selected_pathways) * 0.25),
      units = "in", res = 300)
  do.call(pheatmap::pheatmap, heatmap_args)
  dev.off()
  
  cat("Saved heatmap for", collection_name, value, "\n")
}


#--------genes represented in the enriched M7 gene sets-------------------------
  add_pathway_gene_info <- function(fgsea_result, pathways, rank_list, padj_cutoff = 0.05) {
  
  result_out <- fgsea_result %>%
    as.data.frame() %>%
    filter(padj < padj_cutoff) %>%
    arrange(padj) %>%
    mutate(
      direction = ifelse(NES > 0, "Positive", "Negative"),
      
      # all pathway genes that are present in your ranked gene list
      pathway_genes_in_data = sapply(pathway, function(p) {
        genes <- intersect(pathways[[p]], names(rank_list))
        paste(genes, collapse = ";")
      }),
      
      n_pathway_genes_in_data = sapply(pathway, function(p) {
        length(intersect(pathways[[p]], names(rank_list)))
      }),
      
      # leading-edge genes from fgsea
      leadingEdge = sapply(leadingEdge, paste, collapse = ";"),
      n_leadingEdge = ifelse(
        leadingEdge == "",
        0,
        lengths(strsplit(leadingEdge, ";"))
      )
    ) %>%
    dplyr::select(
      pathway,
      direction,
      NES,
      padj,
      size,
      n_pathway_genes_in_data,
      pathway_genes_in_data,
      n_leadingEdge,
      leadingEdge
    )
  
  return(result_out)
}
```


```{r}
output_dir <- "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/GSEA"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Day 4
fgsea_m7_d4       <- run_fgsea_analysis(rank_d4,        pathways_m7,           "M7",       "Day4",  output_dir)
fgsea_m8_d4       <- run_fgsea_analysis(rank_d4,        pathways_m8,           "M8",       "Day4",  output_dir)
fgsea_kegg_d4     <- run_fgsea_analysis(rank_d4_entrez, pathways_kegg_entrez,  "KEGG",     "Day4",  output_dir)
fgsea_reactome_d4 <- run_fgsea_analysis(rank_d4,        pathways_reactome,     "Reactome", "Day4",  output_dir)

# Day 6
fgsea_m7_d6       <- run_fgsea_analysis(rank_d6,        pathways_m7,           "M7",       "Day6",  output_dir)
fgsea_m8_d6       <- run_fgsea_analysis(rank_d6,        pathways_m8,           "M8",       "Day6",  output_dir)
fgsea_kegg_d6     <- run_fgsea_analysis(rank_d6_entrez, pathways_kegg_entrez,  "KEGG",     "Day6",  output_dir)
fgsea_reactome_d6 <- run_fgsea_analysis(rank_d6,        pathways_reactome,     "Reactome", "Day6",  output_dir)

# Day 11
fgsea_m7_d11       <- run_fgsea_analysis(rank_d11,        pathways_m7,          "M7",       "Day11", output_dir)
fgsea_m8_d11       <- run_fgsea_analysis(rank_d11,        pathways_m8,          "M8",       "Day11", output_dir)
fgsea_kegg_d11     <- run_fgsea_analysis(rank_d11_entrez, pathways_kegg_entrez, "KEGG",     "Day11", output_dir)
fgsea_reactome_d11 <- run_fgsea_analysis(rank_d11,        pathways_reactome,    "Reactome", "Day11", output_dir)

# ── Extract top pathways for plotting ──────────────────────────────────
top_m7_d4  <- get_top_pathways(fgsea_m7_d4)
top_m8_d4  <- get_top_pathways(fgsea_m8_d4)
top_m7_d6  <- get_top_pathways(fgsea_m7_d6)
top_m8_d6  <- get_top_pathways(fgsea_m8_d6)
top_m7_d11 <- get_top_pathways(fgsea_m7_d11)
top_m8_d11 <- get_top_pathways(fgsea_m8_d11)

top_kegg_d4      <- get_top_pathways(fgsea_kegg_d4)
top_kegg_d6      <- get_top_pathways(fgsea_kegg_d6)
top_kegg_d11     <- get_top_pathways(fgsea_kegg_d11)

top_reactome_d4  <- get_top_pathways(fgsea_reactome_d4)
top_reactome_d6  <- get_top_pathways(fgsea_reactome_d6)
top_reactome_d11 <- get_top_pathways(fgsea_reactome_d11)

# Access like:
top_m7_d4$up
top_m7_d4$down


# ── Generate all plots ─────────────────────────────────────────────────
plot_dir <- file.path(output_dir, "plots")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

# Day 4
plot_top_pathways(top_m7_d4,       "M7",       "Day4",  plot_dir)
plot_top_pathways(top_m8_d4,       "M8",       "Day4",  plot_dir)
plot_top_pathways(top_kegg_d4,     "KEGG",     "Day4",  plot_dir)
plot_top_pathways(top_reactome_d4, "Reactome", "Day4",  plot_dir)

# Day 6
plot_top_pathways(top_m7_d6,       "M7",       "Day6",  plot_dir)
plot_top_pathways(top_m8_d6,       "M8",       "Day6",  plot_dir)
plot_top_pathways(top_kegg_d6,     "KEGG",     "Day6",  plot_dir)
plot_top_pathways(top_reactome_d6, "Reactome", "Day6",  plot_dir)

# Day 11
plot_top_pathways(top_m7_d11,       "M7",       "Day11", plot_dir)
plot_top_pathways(top_m8_d11,       "M8",       "Day11", plot_dir)
plot_top_pathways(top_kegg_d11,     "KEGG",     "Day11", plot_dir)
plot_top_pathways(top_reactome_d11, "Reactome", "Day11", plot_dir)



# Package fgsea results by day for each collection
m7_by_day       <- list(Day4 = fgsea_m7_d4,       Day6 = fgsea_m7_d6,       Day11 = fgsea_m7_d11)
m8_by_day       <- list(Day4 = fgsea_m8_d4,       Day6 = fgsea_m8_d6,       Day11 = fgsea_m8_d11)
kegg_by_day     <- list(Day4 = fgsea_kegg_d4,     Day6 = fgsea_kegg_d6,     Day11 = fgsea_kegg_d11)
reactome_by_day <- list(Day4 = fgsea_reactome_d4, Day6 = fgsea_reactome_d6, Day11 = fgsea_reactome_d11)

# NES heatmaps
plot_gsea_heatmap(m7_by_day,       "M7",       plot_dir, value = "NES")
plot_gsea_heatmap(m8_by_day,       "M8",       plot_dir, value = "NES")
plot_gsea_heatmap(kegg_by_day,     "KEGG",     plot_dir, value = "NES")
plot_gsea_heatmap(reactome_by_day, "Reactome", plot_dir, value = "NES")

# Significance heatmaps
plot_gsea_heatmap(m7_by_day,       "M7",       plot_dir, value = "log10padj")
plot_gsea_heatmap(m8_by_day,       "M8",       plot_dir, value = "log10padj")
plot_gsea_heatmap(kegg_by_day,     "KEGG",     plot_dir, value = "log10padj")
plot_gsea_heatmap(reactome_by_day, "Reactome", plot_dir, value = "log10padj")

```
```{r}
library(writexl)
m7_d4_genes <- add_pathway_gene_info(
  fgsea_result = fgsea_m7_d4,
  pathways = pathways_m7,
  rank_list = rank_d4
)

m7_d6_genes <- add_pathway_gene_info(
  fgsea_result = fgsea_m7_d6,
  pathways = pathways_m7,
  rank_list = rank_d6
)

m7_d11_genes <- add_pathway_gene_info(
  fgsea_result = fgsea_m7_d11,
  pathways = pathways_m7,
  rank_list = rank_d11
)

write_xlsx(
  list(
    Day4_M7 = m7_d4_genes,
    Day6_M7 = m7_d6_genes,
    Day11_M7 = m7_d11_genes
  ),
  path = file.path(output_dir, "M7_significant_pathways_with_genes.xlsx")
)
```


#GSEA of M7 and M8 on Day 4
```{r}
# More informative overlap check
pct_m7 <- round(sum(names(rank_d4) %in% unique(msig_m7$gene_symbol)) / length(rank_d4) * 100, 1)
pct_m8 <- round(sum(names(rank_d4) %in% unique(msig_m8$gene_symbol)) / length(rank_d4) * 100, 1)

cat("Ranked genes overlapping M7:", pct_m7, "%\n")
cat("Ranked genes overlapping M8:", pct_m8, "%\n")

# Run GSEA
set.seed(42)
fgsea_m7_d4 <- fgsea(
  pathways = pathways_m7,
  stats = rank_d4,
  minSize = 10,
  maxSize = 500
)

set.seed(42)
fgsea_m8_d4 <- fgsea(
  pathways = pathways_m8,
  stats = rank_d4,
  minSize = 10,
  maxSize = 500
)

#Sort by padj values
fgsea_m7_d4 <- fgsea_m7_d4 %>% arrange(padj)
fgsea_m8_d4 <- fgsea_m8_d4 %>% arrange(padj)

#Inspecting significant pathways
# total pathways tested
n_total_m7d4 <- nrow(fgsea_m7_d4)
n_total_m8d4 <- nrow(fgsea_m8_d4)

# significant pathways (BH adjusted)
n_sig_m7d4 <- sum(fgsea_m7_d4$padj < 0.05, na.rm = TRUE)
n_sig_m8d4 <- sum(fgsea_m8_d4$padj < 0.05, na.rm = TRUE)

cat("M7 Day 4:",
    n_sig_m7d4, "/", n_total_m7d4,
    "pathways significant (",
    round(n_sig_m7d4 / n_total_m7d4 * 100, 1), "% )\n")

cat("M8 Day 4:",
    n_sig_m8d4, "/", n_total_m8d4,
    "pathways significant (",
    round(n_sig_m8d4 / n_total_m8d4 * 100, 1), "% )\n")


#Significant only dataset for plotting
fgsea_m7_d4_sig <- fgsea_m7_d4 %>%
  dplyr::filter(padj < 0.05)

fgsea_m8_d4_sig <- fgsea_m8_d4 %>%
  dplyr::filter(padj < 0.05)

# Collapse leadingEdge before selecting/exporting
fgsea_m7_d4_out <- fgsea_m7_d4 %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))

fgsea_m8_d4_out <- fgsea_m8_d4 %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))

fgsea_m7_d4_sig_out <- fgsea_m7_d4_sig %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))

fgsea_m8_d4_sig_out <- fgsea_m8_d4_sig %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))

# Top 10 positively and negatively enriched pathways M7
top_m7_up <- fgsea_m7_d4_sig_out %>%
  filter(NES > 0) %>%
  arrange(desc(NES)) %>%
  select(pathway, NES, ES, pval, padj, size, leadingEdge) %>%
  slice_head(n = 10)

top_m7_down <- fgsea_m7_d4_sig_out %>%
  filter(NES < 0) %>%
  arrange(NES) %>%
  select(pathway, NES, ES, pval, padj, size, leadingEdge) %>%
  slice_head(n = 10)

# Top 10 positively and negatively enriched pathways M8
top_m8_up <- fgsea_m8_d4_sig_out %>%
  filter(NES > 0) %>%
  arrange(desc(NES)) %>%
  select(pathway, NES, ES, pval, padj, size, leadingEdge) %>%
  slice_head(n = 10)

top_m8_down <- fgsea_m8_d4_sig_out %>%
  filter(NES < 0) %>%
  arrange(NES) %>%
  select(pathway, NES, ES, pval, padj, size, leadingEdge) %>%
  slice_head(n = 10)

# Output only significant pathways
write.csv(
  fgsea_m7_d4_sig_out[, c("pathway", "NES", "ES", "pval", "padj", "size", "leadingEdge")],
  file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m7_day4_sig.csv",
  row.names = FALSE
)

write.csv(
  fgsea_m8_d4_sig_out[, c("pathway", "NES", "ES", "pval", "padj", "size", "leadingEdge")],
  file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m8_day4_sig.csv",
  row.names = FALSE
)

# Output all pathways
write.csv(
  fgsea_m7_d4_out[, c("pathway", "NES", "ES", "pval", "padj", "size", "leadingEdge")],
  file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m7_day4_clean.csv",
  row.names = FALSE
)

write.csv(
  fgsea_m8_d4_out[, c("pathway", "NES", "ES", "pval", "padj", "size", "leadingEdge")],
  file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m8_day4_clean.csv",
  row.names = FALSE
)
```

#Figures for GSEA results of M7 and M8 on day 4
```{r}
#Prepare dataset
plot_m7 <- bind_rows(top_m7_up, top_m7_down) %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    pathway = reorder(pathway, NES)
  )

plot_m8 <- bind_rows(top_m8_up, top_m8_down) %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    pathway = reorder(pathway, NES)
  )

#Barplots
ggplot(plot_m7, aes(x = pathway, y = NES, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top enriched M7 pathways at Day 4",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
  theme_bw()


ggplot(plot_m8, aes(x = pathway, y = NES, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top enriched M8 pathways at Day 4",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
  theme_bw()

#Dotplots
ggplot(plot_m7, aes(x = NES, y = pathway)) +
  geom_point(aes(size = size, color = padj)) +
  scale_color_gradient(low = "darkblue", high = "lightblue") +
  labs(
    title = "Top enriched M7 pathways at Day 4",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    size = "Gene set size",
    color = "Adjusted p-value"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 8),   # pathway names can get cramped
    legend.position = "right",
    plot.title = element_text(face = "bold")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50")  # reference line at NES = 0

ggplot(plot_m8, aes(x = NES, y = pathway)) +
  geom_point(aes(size = size, color = padj)) +
  scale_color_gradient(low = "darkblue", high = "lightblue") +
  labs(
    title = "Top enriched M8 pathways at Day 4",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    size = "Gene set size",
    color = "Adjusted p-value"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 8),   # pathway names can get cramped
    legend.position = "right",
    plot.title = element_text(face = "bold")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50")  # reference line at NES = 0


```


#GSEA of M7 and M8 on Day 6
```{r}
#extract stat
res_d6 <- results(dds6, contrast = c("group", "Vaccinated", "Naive"))

# ranked list for GSEA
rank_d6 <- res_d6$stat
names(rank_d6) <- rownames(res_d6)

# remove NA and sort decreasing
rank_d6 <- rank_d6[!is.na(rank_d6)]
rank_d6 <- sort(rank_d6, decreasing = TRUE)

#Make sure gene names match MSigDB format
head(rownames(dds6))
#Check for duplicated gene names
sum(duplicated(names(rank_d6)))
#Check for infinite values
sum(!is.finite(rank_d6))

#Check if the ranked genes overlap with the gene sets
sum(names(rank_d6) %in% unique(msig_m7$gene_symbol))
sum(names(rank_d6) %in% unique(msig_m8$gene_symbol))

#Run GSEA
fgsea_m7_d6 <- fgsea(
  pathways = pathways_m7,
  stats = rank_d6,
  minSize = 10,
  maxSize = 500
)

fgsea_m8_d6 <- fgsea(
  pathways = pathways_m8,
  stats = rank_d6,
  minSize = 10,
  maxSize = 500
)


#Sort by padj values
fgsea_m7_d6 <- fgsea_m7_d6 %>% arrange(padj)
fgsea_m8_d6 <- fgsea_m8_d6 %>% arrange(padj)

#Inspecting significant pathways
# total pathways tested
n_total_m7d6 <- nrow(fgsea_m7_d6)
n_total_m8d6 <- nrow(fgsea_m8_d6)

# significant pathways (BH adjusted)
n_sig_m7d6 <- sum(fgsea_m7_d6$padj < 0.05, na.rm = TRUE)
n_sig_m8d6 <- sum(fgsea_m8_d6$padj < 0.05, na.rm = TRUE)

# NOT significant pathways
n_not_sig_m7d6 <- sum(fgsea_m7_d6$padj >= 0.05 | is.na(fgsea_m7_d6$padj))
n_not_sig_m8d6 <- sum(fgsea_m8_d6$padj >= 0.05 | is.na(fgsea_m8_d6$padj))

n_total_m7d6
n_sig_m7d6
n_not_sig_m7d6
cat("M7 Day 6:",
    n_sig_m7d6, "/", n_total_m7d6,
    "pathways significant (",
    round(n_sig_m7d6 / n_total_m7d6 * 100, 1), "% )\n")

n_total_m8d6
n_sig_m8d6
n_not_sig_m8d6
cat("M8 Day 6:",
    n_sig_m8d6, "/", n_total_m8d6,
    "pathways significant (",
    round(n_sig_m8d6 / n_total_m8d6 * 100, 1), "% )\n")


#Significant only dataset for plotting
fgsea_m7_d6_sig <- fgsea_m7_d6 %>%
  dplyr::filter(padj < 0.05)

fgsea_m8_d6_sig <- fgsea_m8_d6 %>%
  dplyr::filter(padj < 0.05)


# Top 10 positively and negatively enriched pathways M7
top_m7d6_up <- fgsea_m7_d6_sig %>%
  dplyr::filter(NES > 0) %>%
  dplyr::arrange(desc(NES)) %>%
  dplyr::select(pathway, NES, padj, size, leadingEdge) %>%
  dplyr::slice_head(n = 10)

top_m7d6_down <- fgsea_m7_d6_sig %>%
  dplyr::filter(NES < 0) %>%
  dplyr::arrange(NES) %>%
  dplyr::select(pathway, NES, padj, size, leadingEdge) %>%
  dplyr::slice_head(n = 10)

# Top 10 positively and negatively enriched pathways M8
top_m8d6_up <- fgsea_m8_d6_sig %>%
  dplyr::filter(NES > 0) %>%
  dplyr::arrange(desc(NES)) %>%
  dplyr::select(pathway, NES, padj, size, leadingEdge) %>%
  dplyr::slice_head(n = 10)

top_m8d6_down <- fgsea_m8_d6_sig %>%
  dplyr::filter(NES < 0) %>%
  dplyr::arrange(NES) %>%
  dplyr::select(pathway, NES, padj, size, leadingEdge) %>%
  dplyr::slice_head(n = 10)


#Output only significant
write.csv(fgsea_m7_d6_sig[, c("pathway", "NES", "padj", "size")],
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m7_day6_sig.csv",
          row.names = FALSE)

write.csv(fgsea_m8_d6_sig[, c("pathway", "NES", "padj", "size")],
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m8_day6_sig.csv",
          row.names = FALSE)



write.csv(fgsea_m7_d6[, c("pathway", "NES", "padj", "size")],
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m7_day6_clean.csv",
          row.names = FALSE)

write.csv(fgsea_m8_d6[, c("pathway", "NES", "padj", "size")],
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m8_day6_clean.csv",
          row.names = FALSE)


plotEnrichment(pathways_m7[["CUI_T_CELL_CD8_IL18_RESPONSE_UP"]], rank_d6)
```



#Figures of M7 and M8 on Day 6
```{r}

#Prepare dataset
plot_m7d6 <- bind_rows(top_m7d6_up, top_m7d6_down) %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    pathway = reorder(pathway, NES)
  )

plot_m8d6 <- bind_rows(top_m8d6_up, top_m8d6_down) %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    pathway = reorder(pathway, NES)
  )

#Barplots
ggplot(plot_m7d6, aes(x = pathway, y = NES, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top enriched M7 pathways at Day 6",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
  theme_bw()


ggplot(plot_m8d6, aes(x = pathway, y = NES, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top enriched M8 pathways at Day 6",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
  theme_bw()

#Dotplots
ggplot(plot_m7d6, aes(x = NES, y = pathway)) +
  geom_point(aes(size = size, color = padj)) +
  scale_color_gradient(low = "tomato", high = "steelblue", trans = "reverse") +
  labs(
    title = "Top enriched M7 pathways at Day 6",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    size = "Gene set size",
    color = "Adjusted p-value"
  ) +
  theme_bw()

ggplot(plot_m8d6, aes(x = NES, y = pathway)) +
  geom_point(aes(size = size, color = padj)) +
  scale_color_gradient(low = "tomato", high = "steelblue", trans = "reverse") +
  labs(
    title = "Top enriched M8 pathways at Day 6",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    size = "Gene set size",
    color = "Adjusted p-value"
  ) +
  theme_bw()
```




#GSEA of M7 and M8 on Day 11
```{r}

#Run GSEA
fgsea_m7_d11 <- fgsea(
  pathways = pathways_m7,
  stats = rank_d11,
  minSize = 10,
  maxSize = 500
)

fgsea_m8_d11 <- fgsea(
  pathways = pathways_m8,
  stats = rank_d11,
  minSize = 10,
  maxSize = 500
)


#Sort and inspect
fgsea_m7_d11 <- fgsea_m7_d11 %>% arrange(padj)
fgsea_m8_d11 <- fgsea_m8_d11 %>% arrange(padj)



#Inspecting significant pathways
# total pathways tested
n_total_m7d11 <- nrow(fgsea_m7_d11)
n_total_m8d11 <- nrow(fgsea_m8_d11)

# significant pathways (BH adjusted)
n_sig_m7d11 <- sum(fgsea_m7_d11$padj < 0.05, na.rm = TRUE)
n_sig_m8d11 <- sum(fgsea_m8_d11$padj < 0.05, na.rm = TRUE)

# NOT significant pathways
n_not_sig_m7d11 <- sum(fgsea_m7_d11$padj >= 0.05 | is.na(fgsea_m7_d11$padj))
n_not_sig_m8d11 <- sum(fgsea_m8_d11$padj >= 0.05 | is.na(fgsea_m8_d11$padj))

n_total_m7d11
n_sig_m7d11
n_not_sig_m7d11
cat("M7 Day 11:",
    n_sig_m7d11, "/", n_total_m7d11,
    "pathways significant (",
    round(n_sig_m7d11 / n_total_m7d11 * 100, 1), "% )\n")

n_total_m8d11
n_sig_m8d11
n_not_sig_m8d11
cat("M8 Day 11:",
    n_sig_m8d11, "/", n_total_m8d11,
    "pathways significant (",
    round(n_sig_m8d11 / n_total_m8d11 * 100, 1), "% )\n")


#Significant only dataset for plotting
fgsea_m7_d11_sig <- fgsea_m7_d11 %>%
  dplyr::filter(padj < 0.05)

fgsea_m8_d11_sig <- fgsea_m8_d11 %>%
  dplyr::filter(padj < 0.05)


# Top 10 positively and negatively enriched pathways M7
top_m7d11_up <- fgsea_m7_d11_sig %>%
  dplyr::filter(NES > 0) %>%
  dplyr::arrange(desc(NES)) %>%
  dplyr::select(pathway, NES, padj, size, leadingEdge) %>%
  dplyr::slice_head(n = 10)

top_m7d11_down <- fgsea_m7_d11_sig %>%
  dplyr::filter(NES < 0) %>%
  dplyr::arrange(NES) %>%
  dplyr::select(pathway, NES, padj, size, leadingEdge) %>%
  dplyr::slice_head(n = 10)

# Top 10 positively and negatively enriched pathways M8
top_m8d11_up <- fgsea_m8_d11_sig %>%
  dplyr::filter(NES > 0) %>%
  dplyr::arrange(desc(NES)) %>%
  dplyr::select(pathway, NES, padj, size, leadingEdge) %>%
  dplyr::slice_head(n = 10)

top_m8d11_down <- fgsea_m8_d11_sig %>%
  dplyr::filter(NES < 0) %>%
  dplyr::arrange(NES) %>%
  dplyr::select(pathway, NES, padj, size, leadingEdge) %>%
  dplyr::slice_head(n = 10)


#Output only significant
write.csv(fgsea_m7_d11_sig[, c("pathway", "NES", "padj", "size")],
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m7_day11_sig.csv",
          row.names = FALSE)

write.csv(fgsea_m8_d11_sig[, c("pathway", "NES", "padj", "size")],
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m8_day11_sig.csv",
          row.names = FALSE)


write.csv(fgsea_m7_d11[, c("pathway", "NES", "padj", "size")],
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m7_day11_clean.csv",
          row.names = FALSE)

write.csv(fgsea_m8_d11[, c("pathway", "NES", "padj", "size")],
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/fgsea_m8_day11_clean.csv",
          row.names = FALSE)



```



#Figures for GSEA results of M7 and M8 on Day 11
```{r}
#Prepare dataset
plot_m7d11 <- bind_rows(top_m7d11_up, top_m7d11_down) %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    pathway = reorder(pathway, NES)
  )

plot_m8d11 <- bind_rows(top_m8d11_up, top_m8d11_down) %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    pathway = reorder(pathway, NES)
  )

#Barplots
ggplot(plot_m7d11, aes(x = pathway, y = NES, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top enriched M7 pathways at Day 11",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
  theme_bw()


ggplot(plot_m8d11, aes(x = pathway, y = NES, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top enriched M8 pathways at Day 11",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
  theme_bw()

#Dotplots
ggplot(plot_m7d11, aes(x = NES, y = pathway)) +
  geom_point(aes(size = size, color = padj)) +
  scale_color_gradient(low = "tomato", high = "steelblue", trans = "reverse") +
  labs(
    title = "Top enriched M7 pathways at Day 11",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    size = "Gene set size",
    color = "Adjusted p-value"
  ) +
  theme_bw()

ggplot(plot_m8d11, aes(x = NES, y = pathway)) +
  geom_point(aes(size = size, color = padj)) +
  scale_color_gradient(low = "tomato", high = "steelblue", trans = "reverse") +
  labs(
    title = "Top enriched M8 pathways at Day 11",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    size = "Gene set size",
    color = "Adjusted p-value"
  ) +
  theme_bw()
```


#KEGG
```{r message=FALSE, warning=FALSE}
library(clusterProfiler)
library(org.Mm.eg.db)
library(KEGGREST)
```

#KEGG gene list
```{r message=FALSE, warning=FALSE}
# Step 1: Get direct pathway-gene links (more reliable than keggGet parsing)
pw_gene_raw <- keggLink("mmu", "pathway")

gene_pathway_df <- data.frame(
  entrez  = sub("mmu:", "", pw_gene_raw),          # values are genes
  pathway = sub("path:", "", names(pw_gene_raw)),   # names are pathways
  stringsAsFactors = FALSE
)

# Step 2: Convert all Entrez IDs to symbols in one batch (faster + fewer dropped)
all_entrez <- unique(gene_pathway_df$entrez)
id_map <- bitr(all_entrez, fromType = "ENTREZID", toType = "SYMBOL", OrgDb = org.Mm.eg.db)

# Step 3: Join and split by pathway
gene_pathway_df <- merge(gene_pathway_df, id_map, by.x = "entrez", by.y = "ENTREZID")

pathways_kegg_symbol <- split(gene_pathway_df$SYMBOL, gene_pathway_df$pathway)
pathways_kegg_symbol <- lapply(pathways_kegg_symbol, unique)
pathways_kegg_symbol <- pathways_kegg_symbol[sapply(pathways_kegg_symbol, length) > 0]
```




#Convert gene names
```{r}
# Day 4
# Your ranked list uses gene symbols like "Actb", "Cd8a" etc.
head(names(rank_d4))

# Check KEGG pathways also have symbols (not Entrez IDs)
head(pathways_kegg_symbol[[1]])

# Check overlap — should be reasonably high (>40%)
overlap <- sum(names(rank_d4) %in% unique(unlist(pathways_kegg_symbol)))
cat("Overlapping genes:", overlap, "/", length(rank_d4), 
    "(", round(overlap/length(rank_d4)*100, 1), "%)\n")

#KEGG
set.seed(42)
fgsea_kegg_d4 <- fgsea(
  pathways = pathways_kegg_symbol,
  stats    = rank_d4,
  minSize  = 10,
  maxSize  = 500
)

fgsea_kegg_d4 <- fgsea_kegg_d4 %>% arrange(padj)

# Summary
n_sig_kegg <- sum(fgsea_kegg_d4$padj < 0.05, na.rm = TRUE)
cat("KEGG Day 4:", n_sig_kegg, "/", nrow(fgsea_kegg_d4), "pathways significant\n")

write.csv(kegg_d4_df,
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/kegg_day4_clean.csv",
          row.names = FALSE)

write.csv(top10_kegg_d4,
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/kegg_day4_top10.csv",
          row.names = FALSE)

top10_kegg_d4


# Day 6
gene_df_d6 <- bitr(names(rank_d6),
                   fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Mm.eg.db)

rank_kegg_d6 <- rank_d6[gene_df_d6$SYMBOL]
names(rank_kegg_d6) <- gene_df_d6$ENTREZID
rank_kegg_d6 <- sort(rank_kegg_d6, decreasing = TRUE)

gsea_kegg_d6 <- gseKEGG(
  geneList = rank_kegg_d6,
  organism = "mmu",
  minGSSize = 10,
  pvalueCutoff = 0.05,
  eps = 0,
  pAdjustMethod = "BH"
)

kegg_d6_df <- as.data.frame(gsea_kegg_d6) %>%
  dplyr::arrange(p.adjust) %>%
  dplyr::select(Description, NES, p.adjust, setSize)

top10_kegg_d6 <- kegg_d6_df %>%
  dplyr::slice_head(n = 10)

write.csv(kegg_d6_df,
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/kegg_day6_clean.csv",
          row.names = FALSE)

write.csv(top10_kegg_d6,
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/kegg_day6_top10.csv",
          row.names = FALSE)

top10_kegg_d6


# Day 11
gene_df_d11 <- bitr(names(rank_d11),
                    fromType = "SYMBOL",
                    toType = "ENTREZID",
                    OrgDb = org.Mm.eg.db)

rank_kegg_d11 <- rank_d11[gene_df_d11$SYMBOL]
names(rank_kegg_d11) <- gene_df_d11$ENTREZID
rank_kegg_d11 <- sort(rank_kegg_d11, decreasing = TRUE)

gsea_kegg_d11 <- gseKEGG(
  geneList = rank_kegg_d11,
  organism = "mmu",
  minGSSize = 10,
  pvalueCutoff = 0.05,
  eps = 0,
  pAdjustMethod = "BH"
)


kegg_d11_df <- as.data.frame(gsea_kegg_d11) %>%
  dplyr::arrange(p.adjust) %>%
  dplyr::select(Description, NES, p.adjust, setSize)

top10_kegg_d11 <- kegg_d11_df %>%
  dplyr::slice_head(n = 10)

write.csv(kegg_d11_df,
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/kegg_day11_clean.csv",
          row.names = FALSE)

write.csv(top10_kegg_d11,
          file = "/work/users/x/c/xczhou/Chlamydia/Analysis Outputs/kegg_day11_top10.csv",
          row.names = FALSE)

top10_kegg_d11
```

#Top pathway tables for each day
```{r}
# Day 4
top_kegg_d4_up <- kegg_d4_df %>%
  dplyr::filter(NES > 0) %>%
  dplyr::arrange(desc(NES)) %>%
  dplyr::slice_head(n = 10)

top_kegg_d4_down <- kegg_d4_df %>%
  dplyr::filter(NES < 0) %>%
  dplyr::arrange(NES) %>%
  dplyr::slice_head(n = 10)

# Day 6
top_kegg_d6_up <- kegg_d6_df %>%
  dplyr::filter(NES > 0) %>%
  dplyr::arrange(desc(NES)) %>%
  dplyr::slice_head(n = 10)

top_kegg_d6_down <- kegg_d6_df %>%
  dplyr::filter(NES < 0) %>%
  dplyr::arrange(NES) %>%
  dplyr::slice_head(n = 10)

# Day 11
top_kegg_d11_up <- kegg_d11_df %>%
  dplyr::filter(NES > 0) %>%
  dplyr::arrange(desc(NES)) %>%
  dplyr::slice_head(n = 10)

top_kegg_d11_down <- kegg_d11_df %>%
  dplyr::filter(NES < 0) %>%
  dplyr::arrange(NES) %>%
  dplyr::slice_head(n = 10)


#Prepare plotting datasets
plot_kegg_d4 <- bind_rows(top_kegg_d4_up, top_kegg_d4_down) %>%
  dplyr::mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    Description = reorder(Description, NES)
  )

plot_kegg_d6 <- bind_rows(top_kegg_d6_up, top_kegg_d6_down) %>%
  dplyr::mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    Description = reorder(Description, NES)
  )

plot_kegg_d11 <- bind_rows(top_kegg_d11_up, top_kegg_d11_down) %>%
  dplyr::mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    Description = reorder(Description, NES)
  )
```




#Visualizations of KEGG Barplots
```{r}
# Day 4 bar plot
ggplot(plot_kegg_d4, aes(x = Description, y = NES, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top enriched KEGG pathways at Day 4",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
  theme_bw()

# Day 6 bar plot
ggplot(plot_kegg_d6, aes(x = Description, y = NES, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top enriched KEGG pathways at Day 6",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
  theme_bw()

# Day 11 bar plot
ggplot(plot_kegg_d11, aes(x = Description, y = NES, fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top enriched KEGG pathways at Day 11",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "tomato")) +
  theme_bw()
```



#Visualizations of KEGG Dotplots
```{r}
# Day 4 dot plot
ggplot(plot_kegg_d4, aes(x = NES, y = Description)) +
  geom_point(aes(size = setSize, color = p.adjust)) +
  scale_color_gradient(low = "tomato", high = "steelblue", trans = "reverse") +
  labs(
    title = "Top enriched KEGG pathways at Day 4",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    size = "Gene set size",
    color = "Adjusted p-value"
  ) +
  theme_bw()

# Day 6 dot plot
ggplot(plot_kegg_d6, aes(x = NES, y = Description)) +
  geom_point(aes(size = setSize, color = p.adjust)) +
  scale_color_gradient(low = "tomato", high = "steelblue", trans = "reverse") +
  labs(
    title = "Top enriched KEGG pathways at Day 6",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    size = "Gene set size",
    color = "Adjusted p-value"
  ) +
  theme_bw()

# Day 11 dot plot
ggplot(plot_kegg_d11, aes(x = NES, y = Description)) +
  geom_point(aes(size = setSize, color = p.adjust)) +
  scale_color_gradient(low = "tomato", high = "steelblue", trans = "reverse") +
  labs(
    title = "Top enriched KEGG pathways at Day 11",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    size = "Gene set size",
    color = "Adjusted p-value"
  ) +
  theme_bw()
```



#Reactome day 4
```{r}
fgsea_reactome_d4 <- fgsea(
  pathways = pathways_reactome,
  stats = rank_d4,
  minSize = 10,
  maxSize = 500
)

fgsea_reactome_d4 <- fgsea_reactome_d4 %>%
  arrange(padj)


#Prepare Reactome results for plotting
#Removes "REACTOME_" from the beginning
#Replaces underscores with space
reactome_d4_df <- fgsea_reactome_d4 %>%
  as.data.frame() %>%
  filter(!is.na(padj)) %>%
  mutate(
    Description = str_replace_all(pathway, "^REACTOME_", ""),
    Description = str_replace_all(Description, "_", " ")
  )

#Significant Pathways only
reactome_d4_sig <- reactome_d4_df %>%
  filter(padj < 0.05) %>%
  arrange(desc(NES))

get_top10 <- function(df, direction) {
  if (direction == "up") {
    df %>%
      filter(NES > 0) %>%
      arrange(desc(NES)) %>%
      slice_head(n = 10)
  } else {
    df %>%
      filter(NES < 0) %>%
      arrange(NES) %>%
      slice_head(n = 10)
  }
}

reactome_d4_top <- bind_rows(
  get_top10(reactome_d4_sig, "up"),
  get_top10(reactome_d4_sig, "down")
) %>%
  arrange(NES)

#Bar plot
ggplot(reactome_d4_top,
       aes(x = reorder(Description, NES), y = NES)) +
  geom_col(aes(fill = NES > 0)) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  coord_flip() +
  labs(x = "Reactome pathway",
       y = "Normalized Enrichment Score (NES)",
       title = "Significant Reactome pathways at Day 4") +
  theme_bw() +
  theme(legend.position = "none")

#Dotplot
ggplot(reactome_d4_top,
       aes(x = NES,
           y = reorder(Description, NES),
           size = -log10(padj),
           color = NES)) +
  geom_point() +
  labs(x = "Normalized Enrichment Score (NES)",
       y = "Reactome pathway",
       title = "Significant Reactome pathways at Day 4",
       size = "-log10 adjusted p-value") +
  theme_bw()



```

#Reactome day6
```{r}
fgsea_reactome_d6 <- fgsea(
  pathways = pathways_reactome,
  stats = rank_d6,
  minSize = 10,
  maxSize = 500
)

fgsea_reactome_d6 <- fgsea_reactome_d6 %>%
  arrange(padj)

# Prepare Reactome results for plotting
reactome_d6_df <- fgsea_reactome_d6 %>%
  as.data.frame() %>%
  filter(!is.na(padj)) %>%
  mutate(
    Description = str_replace_all(pathway, "^REACTOME_", ""),
    Description = str_replace_all(Description, "_", " ")
  )

# Significant pathways only
reactome_d6_sig <- reactome_d6_df %>%
  filter(padj < 0.05) %>%
  arrange(desc(NES))

reactome_d6_top <- bind_rows(
  get_top10(reactome_d6_sig, "up"),
  get_top10(reactome_d6_sig, "down")
) %>%
  arrange(NES)

# Bar plot
ggplot(reactome_d6_top,
       aes(x = reorder(Description, NES), y = NES)) +
  geom_col(aes(fill = NES > 0)) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  coord_flip() +
  labs(x = "Reactome pathway",
       y = "Normalized Enrichment Score (NES)",
       title = "Significant Reactome pathways at Day 6") +
  theme_bw() +
  theme(legend.position = "none")

# Dot plot
ggplot(reactome_d6_top,
       aes(x = NES,
           y = reorder(Description, NES),
           size = -log10(padj),
           color = NES)) +
  geom_point() +
  labs(x = "Normalized Enrichment Score (NES)",
       y = "Reactome pathway",
       title = "Significant Reactome pathways at Day 6",
       size = "-log10 adjusted p-value") +
  theme_bw()
```


#Reactome day 11
```{r}
# Reactome day 11

fgsea_reactome_d11 <- fgsea(
  pathways = pathways_reactome,
  stats = rank_d11,
  minSize = 10,
  maxSize = 500
)

fgsea_reactome_d11 <- fgsea_reactome_d11 %>%
  arrange(padj)

# Prepare Reactome results for plotting
reactome_d11_df <- fgsea_reactome_d11 %>%
  as.data.frame() %>%
  filter(!is.na(padj)) %>%
  mutate(
    Description = str_replace_all(pathway, "^REACTOME_", ""),
    Description = str_replace_all(Description, "_", " ")
  )

# Significant pathways only
reactome_d11_sig <- reactome_d11_df %>%
  filter(padj < 0.05) %>%
  arrange(desc(NES))

reactome_d11_top <- bind_rows(
  get_top10(reactome_d11_sig, "up"),
  get_top10(reactome_d11_sig, "down")
) %>%
  arrange(NES)


# Bar plot
ggplot(reactome_d11_top,
       aes(x = reorder(Description, NES), y = NES)) +
  geom_col(aes(fill = NES > 0)) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  coord_flip() +
  labs(x = "Reactome pathway",
       y = "Normalized Enrichment Score (NES)",
       title = "Significant Reactome pathways at Day 11") +
  theme_bw() +
  theme(legend.position = "none")

# Dot plot
ggplot(reactome_d11_top,
       aes(x = NES,
           y = reorder(Description, NES),
           size = -log10(padj),
           color = NES)) +
  geom_point() +
  labs(x = "Normalized Enrichment Score (NES)",
       y = "Reactome pathway",
       title = "Significant Reactome pathways at Day 11",
       size = "-log10 adjusted p-value") +
  theme_bw()
```


#Overlapping genes
```{r}
# Get all unique genes in each collection
m7_genes <- unique(unlist(pathways_m7))
m8_genes <- unique(unlist(pathways_m8))
reactome_genes <- unique(unlist(pathways_reactome))
kegg_genes <- unique(unlist(pathways_kegg_symbol))

# Pairwise overlaps
overlap_summary <- rbind(
  calc_overlap(m7_genes, m8_genes,       "M7",      "M8"),
  calc_overlap(m7_genes, reactome_genes, "M7",      "Reactome"),
  calc_overlap(m7_genes, kegg_genes,     "M7",      "KEGG"),
  calc_overlap(m8_genes, reactome_genes, "M8",      "Reactome"),
  calc_overlap(m8_genes, kegg_genes,     "M8",      "KEGG"),
  calc_overlap(reactome_genes, kegg_genes, "Reactome", "KEGG")
)
overlap_summary

# All 4-way overlap
overlap_all4 <- Reduce(intersect, list(m7_genes, m8_genes, reactome_genes, kegg_genes))

all4_summary <- data.frame(
  comparison      = "M7 vs M8 vs Reactome vs KEGG",
  n_overlap       = length(overlap_all4),
  out_of_M7       = length(m7_genes),
  pct_M7          = round(length(overlap_all4) / length(m7_genes) * 100, 2),
  out_of_M8       = length(m8_genes),
  pct_M8          = round(length(overlap_all4) / length(m8_genes) * 100, 2),
  out_of_Reactome = length(reactome_genes),
  pct_Reactome    = round(length(overlap_all4) / length(reactome_genes) * 100, 2),
  out_of_KEGG     = length(kegg_genes),
  pct_KEGG        = round(length(overlap_all4) / length(kegg_genes) * 100, 2)
)
all4_summary

overlap_all4 <- Reduce(
  intersect,
  list(m7_genes, m8_genes, reactome_genes, kegg_genes)
)

all4_summary <- data.frame(
  comparison = "M7 vs M8 vs Reactome vs KEGG",
  n_overlap = length(overlap_all4),
  out_of_M7 = length(m7_genes),
  pct_M7 = round(length(overlap_all4) / length(m7_genes) * 100, 2),
  out_of_M8 = length(m8_genes),
  pct_M8 = round(length(overlap_all4) / length(m8_genes) * 100, 2),
  out_of_Reactome = length(reactome_genes),
  pct_Reactome = round(length(overlap_all4) / length(reactome_genes) * 100, 2),
  out_of_KEGG = length(kegg_genes),
  pct_KEGG = round(length(overlap_all4) / length(kegg_genes) * 100, 2)
)

all4_summary

#Visualization
library(ggVennDiagram)

gene_sets <- list(
  M7       = m7_genes,
  M8       = m8_genes,
  Reactome = reactome_genes,
  KEGG     = kegg_genes
)

ggVennDiagram(gene_sets) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Gene overlap across pathway collections")
```

